---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# SDS230: Data Exploration and Analysis

<!-- badges: start -->
<!-- badges: end -->


Class material and functions for S&DS 230: Data Exploration and Analysis


## Installing the SDS230 package

To install the SDS230 package that contains functions that are useful for the
class run the function below 

Note: we recommend you cut and paste all these commands to avoid typos.


```{r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("emeyers/SDS230")
```


## Initial setup using the SDS230 package

Once you have installed the SDS230 package, you can use this package to install other packaged required by the class as well as LaTeX using the `SDS230:::initial_setup()` function shown below. 

Notes: 

  1. If any dialog boxes pop up, just click "ok" on them to proceed. 
  2. This function might take several minutes to run so please be patient. 
  3. If it asks you to update any existing package you can do so (recommended) or you can skip this.
  4. When the function is done running, if it asks you to restart R, please close and then reopen R Studio. 

```{r, eval=FALSE}
SDS230:::initial_setup()
```


#### Testing LaTeX has been installed 

To test that LaTeX has correctly installed on your system you can run the following command: 

```{r test_tinytex_install}
tinytex:::is_tinytex()
```

If the function returns TRUE, then you have successfully installed LaTeX. 



## Class material

All class material is in the
[ClassMaterial](https://github.com/emeyers/SDS230/tree/master/ClassMaterial)
directory. In particular the
[ClassMaterial/slides](https://github.com/emeyers/SDS230/tree/master/ClassMaterial/slides)
directory contains the class slides which can be useful to review what was
covered in class.



