---
title: "Class 14 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


SDS230::download_data("amazon-books.txt")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(tidyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```






$\\$






## Overview

 * Review mapping and linear regression




$\\$








## Part 1: Review maps and linear regression




#### Part 1.1: maps


Let's create a choropleth map of the world where the color scale indicates how much alcohol is consumed per person. 

We will use `drinks` data frame from the fivethirtyeight package which contains this information. For more information on this data set see: https://fivethirtyeight.com/features/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/

To learn more about what your favorite map projection says about you, see: https://xkcd.com/977/




```{r}

# install.packages("fivethirtyeight")

library(fivethirtyeight)

#glimpse(drinks)


# get a map of the world
head(map_data("world"))


# create a choropleth map of how much alcohol is consumed per capita
map_data("world") %>%
  left_join(drinks, by = c("region" = "country")) %>%
  arrange(group, order) %>%
  ggplot(aes(x = long, y = lat, group = group, fill = total_litres_of_pure_alcohol)) +   
  geom_polygon(color = "black") +
  scale_fill_continuous(type = "viridis") +
  xlab("") + 
  ylab("")  # +  coord_map("ortho")


```


$\\$




#### Part 1.2: linear regression 


To quickly review linear regression, let's build a linear model to predict how popular a given type of halloween candy is based on how much sugar the candy has in it. The data we will use is in the data frame `candy_rankings` from the fivethirtyeight package. The variables we will look at are `sugarpercent` which is the percent of sugar a piece of candy has, and `winpercent`, which is the percent of time people choose a particular type of candy over an alternative type of candy. For more information about the data see: https://fivethirtyeight.com/videos/the-ultimate-halloween-candy-power-ranking/



```{r}


# View the data frame, create a scatter plot between sugarpercent and winpercent 
#  and calculate the correlation 

#glimpse(candy_rankings)

plot(candy_rankings$sugarpercent, candy_rankings$winpercent)

cor(candy_rankings$sugarpercent, candy_rankings$winpercent)


# fit a linear model and add the model line to the scatter plot
(lm_fit <- lm(winpercent ~ sugarpercent, data = candy_rankings))

abline(lm_fit, col = "red")


# get the regression coefficients, can you write an equation for the model in LaTeX? 
coef(lm_fit)



# examine the residuals 

#lm_fit$residuals

#candy_rankings$winpercent - lm_fit$fitted.values 


# calculate the sum squared residuals (SSE)
# this is also called the residual sum of squares (RSS)

sum(lm_fit$residuals^2)


```






$\\$




