---
title: "Class 11 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
#SDS230::update_installed_packages()


# install.packages("gapminder")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)

library(dplyr)


#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```






$\\$






## Overview


 * Plotting with ggplot



$\\$








## Part 1: Visualizations with ggplot


We can use the ggplot2 package, which is part of the tidyverse, to create much
nicer looking graphics than using base R graphics. The ggplot2 library is
modeled on Leland Wilkinson's "grammar of graphics" which creatse
graphics from a combination of basic visual elements.

In the exercises below, we will learn how to use ggplot using the motor trends
cars data set (mtcars) that comes with base R installation, and also the
gapminder data.


A few resources to learn more about ggplot are: 

- [R for Data Science](https://r4ds.had.co.nz/)
- [The ggplot cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)





$\\$





### Part 1.1: Scatter plots


Let's create plots of the number of miles per gallon (mpg) cars get as a
function of the weight of the car.


```{r scatter_plot}


# install.packages("ggplot2")

library(ggplot2)



# base R





# ggplot 






# ggplot - mapping in the geom





```




$\\$





### Part 1.2: Adding labels to plots


We can add labels to the plots using the `labs()` function. Arguments to `labs()` function include:
- `x`: the label on the x-axis (you can also use the `xlab()` function)
- `y`: the label on the y-axis (you can also use the `ylab()` function)
- `title`: the title of the plot  (you can also use the `ggtitle()` function)
- `subtitle`: the title of the plot  (you can also use the `ggtitle()` function)

If you just want to add x and y labels or a title, you can also use the
`xlab("label1")`, `ylab("label2")` and/or the `ggtitle()` functions.



```{r labels}












```



Remember, if you don't want exes [label your axes](https://xkcd.com/833/)



$\\$




### Part 1.3: Adding annotations to plots

We can add annotations to plots using the `annotate("text", x = , y = , label = )` function.



```{r annotations}










```







$\\$





### Part 1.4: More aesthetic mappings


We can use other aesthetic mappings beyond position including:
- `color`: different color scales are used for quantitative and categorical (factor) data 
- `shape`: should be categorical data
- `size`: should be quantitative data



```{r mappings}


# add color based on the transmission type (is automatic or not)




# it is better to treat the transmission type as a categorical variable?





# can also try mapping transmission type to shape or size 









```



Question: When adding the variable automatic/manual transmission (am) to the
scatter plot mpg vs. weight, do you think it is best to map am on to...?

a. color
b. shape
c. size




$\\$





### Part 1.5: Attributes vs. Aesthetics


Setting an **aesthetic mapping** maps a variable to a **glyph property**. This
is done **inside** the `aes()` function.

Setting an **attribute** set a glyph property to a **fixed value**. This is done
**outside** the `aes()` function.



```{r aesthetics_vs_attributes}


# setting an aesthetic mapping




# setting an attribute





```




$\\$





### Part 1.6: Facets


Beyond comparing variables based on aesthetic mappings, you can compare
categorical variables by splitting a plot into subplots, called facets, using
`facet_wrap()`


```{r facets}


# separate subplots for the two transmission types





# One can also do facets in two dimensions





```




$\\$





### Part 1.7: Overplotting


Sometimes points overlap making it hard to estimate the number of points at a
particular range of values.

We can control the transparency of points by changing their alpha values.



```{r transparency}

library(gapminder)


# a lot of overplotting




# changing the transparency levels makes it a bit easier to see how many points are at a given x, y location




```





$\\$





### Part 1.8: Changing scales


Each visual attribute that has an aesthetic mapping has a default scales. We can
change the scales used for each mapping using functions that start with
`scale_`.

For example, we can change the x-scale from liner to logarithmic using
`scale_x_continuous(trans='log10')`. Likewise we can change the color scale
using `scale_color_manual()`.



```{r scales}

# changing the scale on the x-axis





# mapping continents to colors, and adding my own color scale





```






$\\$






### Part 1.9: geoms


We can use different geoms to create other types of plots.



```{r geoms1}


# Let's create a plot that shows the GDP in the United States as a function of the year using the geom geom_line()





# Let's create a plot that shows the GDP in the United States as a function of the year using the geom geom_col()





# Let's MPG as a function of weight using the names of cars rather than just points for each car





# Let's plot a histogram of the weights of cars




# Let's create a boxplot of the weights of cars



# Let's create a side-by-side boxplot of the weights of cars depending on the number of cylinders the engine has




```





$\\$





### Part 1.10: geoms continued

Violin and Joy plots are other ways to view distributions of data



```{r geoms2}

# violin plot 




library("ggridges")

# joy plot




```


Question: Can you figure out where the name "joy plot" comes from? 





$\\$






### Part 1.11: geoms continued


We can also have multiple geom layers on a single graph by using the + symbol
E.g   `ggplot(…) + geom_type1() + geom_type2()`



```{r geoms3}

# Create a scatter plot of miles per gallon as a function of weight and then add a smoothed line using geom_smooth() and a vertical lines using geom_vline()







```





$\\$





### Part 1.12: Themes


We can also use different themes to change the appearance of our plot.


```{r themes}


# Add theme_classic() to our plot



# install.packages("ggthemes")


# library(ggthemes)







```




$\\$




### Part 1.13: Exeriment yourself


Try to create some interesting visualizations from either a data set we have used in the class, or a new data set you found.




```{r experiment}










```






$\\$





