---
title: "Class 4 notes and code"
output:
  pdf_document: default
editor_options: 
  chunk_output_type: console
---



$\\$



<!--  Please run the code in the  R chunk below once. It will install a packages useful for adding LaTeX to figures and download some data we will use.  -->

```{r, eval = FALSE}


# get some data and install a package that is needed
#install.packages("latex2exp")


# download the okcupid data if you don't have it already
download.file("https://raw.githubusercontent.com/emeyers/SDS230/master/ClassMaterial/data/profiles_revised.csv", "profiles_revised.csv", mode = "wb")


```




```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

set.seed(230)

```




$\\$





## Part 0:  For loops


For loops are useful when you want to repeat a piece of code many times under similar conditions

Print the numbers from 1 to 50...


```{r}








```



$\\$




For loops are particular useful in combination with vectors that can store the results.

Create a vector with the squares of the numbers from 1 to 50. 



```{r}



# create a loop that creates a vector with the squares of the numbers from 1 to 50. 


for (i in 1:50) {
  
  
  
}




# plot the results





```






$\\$








### Try this at home!

Use a for loop to create a vector called `the_results` that holds the values at multiples of 3 from 3 to 300; i.e., `the_results` should hold the numbers 3, 6, 9, ..., 300


```{r}








```






$\\$









## Part 1: Probability functions and generating random numbers in R





###  Part 1.1a: Generating random numbers in R

R has built in functions to generate data from different distributions. All these functions start with the letter `r`. 

We can set the random number generator **seed** to always get the same sequence of random numbers.

Let's get a sample of n = 200 random points from the uniform distribution using `runif()`

```{r try_runif}

# set the seed to a specific number to always get the same sequence of random numbers
set.seed(230)


# generate n = 100 points from U(0, 1) using runif() function 



# plot a histogram of these random numbers




```



### Part 1.1b: Generating random numbers in R


There are many other distributions we can get random numbers from including: 

 - Normal distributions: `rnorm()`
 - Uniform distribution `runif()`
 - Exponential distributions: `rexp()`

And many more!


The first argument to all these functions is the number of random points you want to generate (`n`) and then there are additional arguments that can be used to control the shape of the distribution (i.e., that set the "parameters" of the distribution), 



```{r try_other_random_data_functions}


# generate n = 1000 points from standard normal distribution N(0, 1)



# plot a histogram of these random numbers



```




$\\$




### Part 1.1c: Sampling random values from a vector 


We can sample random `n` random values from a vector `v` using the `sample(v, n)` function. 

We can also set the `replace` argument to `TRUE` to sample values with replacement; e.g., to sample with replacement we can use `sample(v, n, replace = TRUE)`.


Let's create a vector of numbers from 1 to 100 and sample 30 of them randomly (i.e., n = 30). 



```{r sample_values}

# set the seed to always get the same results
# in general, best to just do this once at the top of the RMarkdown document
set.seed(230)

# create a vector of values from 1 to 100



# sample 30 random values 



# plot the values sorted using the sort() function




# sample 30 random values with replacement



# plot the values sorted using the sort() function



```







$\\$






## Part 2: Sampling distributions


### Part 2.1

A distribution of statistics is called a **sampling distribution**. 

Can you generate and plot an approximate sampling distribution for:
 * sample means $\bar{x}$'s 
 * sample size n = 100
 * for data that come from uniform distribution


Note the shape of the *sampling distribution* can be quite different from the shape of the data distribution (which is uniform here).


```{r sampling_dist}


# create a sampling distribution of the mean using data from a uniform distribution
set.seed(67)


sampling_dist <- NULL

for (i in 1:10000) {
  

  

}






# plot a histogram of the sampling distribution of these means 





```






$\\$






### Part 2.2: The standard error (SE)

The deviation of a sampling distribution is called the standard error (SE). Can you calculate (an approximate) standard error for the sampling distribution you created above? 

```{r se}





```






$\\$






### Part 2.3: Approximating a sampling distributions from a data set


We generate samples from an actual data set we have using the `sample()` function.  

Let's start by just generate a single sample of size n = 100 from the OkCupid users' heights and calculating the mean of this sample.


```{r okcupid_sampling_dist1}

# read in the okcupid data
profiles <- read.csv("profiles_revised.csv")


# get the heights for the OkCupid data



# get one random sample of heights from 100 people



# get the mean of this sample



```





$\\$





We can then create an approximation of a sampling distribution from the OkCupid users' data set by repeating this many times in a for loop.



```{r okcupid_sampling_dist2}


# repeat the process 1,000 times
sampling_dist <- NULL 






# plot a histogram of this sampling distribution 




```




**Question:** What would have to be true for this to be an actual sampling distribution? 







$\\$






### Part 2.4:  The central limit theorm



The [central limit theorm (CTL)](https://en.wikipedia.org/wiki/Central_limit_theorem) establishes that (in most situations) when independent random variables are added their (normalized) sum converges to a normal distribution.

Put another way, if we define the average random (i.i.d) sample {$X_1$, $X_2$, ..., $X_n$} of size *n* as: 

$S_{n}:={\frac{X_{1}+\cdots +X_{n}}{n}}$

then the CTL tells us that:

$\sqrt{n}(S_{n} - \mu)$ $\xrightarrow {d} N(0,\sigma^{2})$


You will explore this more through simulations on homework 2. 






$\\$

