---
title: "Class 22 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


SDS230::download_data("IPED_salaries_2016.rda")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Logistic regression
 * Visualizing multiple regression models with ggplot



$\\$




## Part 1: Logistic regression


Let us fit a logistic regression model that gives the probability that a professor is a Full professor (compared to an Assistant Professor) based on the professor's salary; i.e., the model is giving P(Full professor | salary). 

(In the IPED data, the cases are actually individual universities, and the salaries are averages over all professors of a particular rank, so what the model is really doing is giving P(Full professor rank | average salary) which is a little different than the variability that would be seen at the individual professor level). 

Let's start by loading the data, doing a little data cleaning, and plotting the data.



```{r}


# load the data into R
load("IPED_salaries_2016.rda")


# get assistant and full professor salaries 
assistant_full_data <- IPED_salaries %>%
  filter(endowment > 0, rank_name %in% c("Assistant", "Full")) %>%
  select(school, salary_tot, endowment, enroll_total, rank_name) %>%
  na.omit() %>%
  mutate(log_endowment = log10(endowment)) %>%
  mutate(rank_name = relevel(rank_name, "Assistant"))


# visualize the data





  
```





$\\$





#### Part 1.1.: Fitting a logistic model to the data


Let's now fit a logistic regression model using the `glm()` function. To fit a logistic regression model, we will use the family = "binomial" argument. What this function is doing is finding the "maximum likelihood" values for $\hat{\beta}_0$ and $\hat{\beta}_1$.

We can use the coefficients from the model to build a function that can give the probability of that the faculty rank is Full professor based on the salary value:

$P(\text{Full professor} | \text{salary}) = \frac{e^{\hat{\beta}_0 + \hat{\beta}_1 \cdot \text{salary}}}{1 + e^{\hat{\beta}_0 + \hat{\beta}_1 \cdot \text{salary}}}$



```{r}


# build the logistic regression function 


# extract the coefficients



# create the prediction function 

  

# what is the probability that if a school is paying $80,000 on average to a 
# particular faculty rank, this rank corresponds to the Full professor rank?
# (instead of the Assistant Professor rank)? 



```





$\\$





#### Part 1.2.: Plotting the logistic regression function 


Let's now plot the fitted logistic regression function to see how the probability that the rank corresponds to Full professor changes as a function of the salary amount. 



```{r, warning = FALSE}



# to plot this function we add 1 to it



# plot the logistic regression function  









```





$\\$






#### Part 1.3: Multiple logistic regression


We can use multiple predictors in our logistic regression function as well. Let's create a function that gives the probability that the rank corresponds to Full professor given a salary amount a log10 endowment size.



```{r}


# fit the model


# extract the regression coefficients




# build a function that can predict faculty rank based on salary and endowment



# predict probability Full professor if average salary is $80,000 
# and the school has a $10 million endowment ($10^7)



# predict probability Full professor if average salary is $80,000 
# and the school has a $10 billion endowment ($10^10)



```





$\\$






#### Part 1.4: Visualzing the multiple logistic regression function


Let's now visualize the probability function P(Full professor | salary, log_endowment)!



```{r}


# create a 2D plot of the probability that a car is new as a function of price and mileage




# there are more efficient ways to do this using apply or mapping functions





# create the plot





```




$\\$





## Part 2: Visualizing multiple regression models with ggplot

In prior classes we have visualized multiple regression models with categorical predictors using base R graphics. In particular, we created scatter plots for data in different categories using the `plot()` and `points()` and used the `col` argument to color the points. We then added on regression lines using the `abline()` function. This method was useful for educational purposes so that we could see the connection between the model parameters that were estimated using the `lm()` function, and the underlying data. However, if we want to create better looking visualizations in a more efficient manner, then it is better to use ggplot.

Let's now use ggplot to visualize a multiple regression model that has one quantitative and one categorical variable. In particular, let's recreate the plot from class 19 part 3.6 where we display faculty salaries as a function of log endowment separately for different faculty ranks. 



```{r}

# create the IPED data subset used in class 19 part 3
IPED_2 <- IPED_salaries %>%
filter(endowment > 0) %>%
mutate(log_endowment = log10(endowment)) %>%
filter(CARNEGIE %in% c(15, 31)) %>%
filter(rank_name %in% c("Assistant", "Associate", "Full"))


# fit the model
fit_prof_rank <- lm(salary_tot ~ log_endowment + rank_name + log_endowment * rank_name, data = IPED_2)


# using base R graphics - as was done in class 19, part 3.6
plot(salary_tot ~ log_endowment, data = filter(IPED_2, rank_name == "Assistant"), ylim = c(0, 250000))
points(salary_tot ~ log_endowment, data = filter(IPED_2, rank_name == "Associate"), col = 'red')
points(salary_tot ~ log_endowment, data = filter(IPED_2, rank_name == "Full"), col = 'blue')

the_coefs <- coef(fit_prof_rank)
abline(the_coefs[1], the_coefs[2], col = "blue")
abline(the_coefs[1] + the_coefs[3], the_coefs[2] + the_coefs[5], col = "red")
abline(the_coefs[1] + the_coefs[4], the_coefs[2] + the_coefs[6], col = "black")



# using ggplot






```








