---
title: "Class 14 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


SDS230::download_data("amazon-books.txt")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(tidyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```






$\\$






## Overview

 * Review mapping and linear regression
 * Reshaping data
 * Hypothesis tests for more than 2 means (if there is time)





$\\$








## Part 1: Review maps and linear regression




#### Part 1.1: maps


Let's create a choropleth map of the world where the color scale indicates how much alcohol is consumed per person. 

We will use `drinks` data frame from the fivethirtyeight package which contains this information. For more information on this data set see: https://fivethirtyeight.com/features/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/

To learn more about what your favorite map projection says about you, see: https://xkcd.com/977/




```{r}

# install.packages("fivethirtyeight")

library(fivethirtyeight)



# get a map of the world


# create a choropleth map of how much alcohol is consumed per capita






```


$\\$




#### Part 1.2: linear regression 


To quickly review linear regression, let's build a linear model to predict how popular a given type of halloween candy is based on how much sugar the candy has in it. The data we will use is in the data frame `candy_rankings` from the fivethirtyeight package. The variables we will look at are `sugarpercent` which is the percent of sugar a piece of candy has, and `winpercent`, which is the percent of time people choose a particular type of candy over an alternative type of candy. For more information about the data see: https://fivethirtyeight.com/videos/the-ultimate-halloween-candy-power-ranking/



```{r}


# View the data frame, create a scatter plot between sugarpercent and winpercent 
#  and calculate the correlation 





# fit a linear model and add the model line to the scatter plot




# get the regression coefficients - can you write an equation for the model in LaTeX? 




# examine the residuals 




# calculate the sum squared residuals (SSE)
# this is also called the residual sum of squares (RSS)




```






$\\$






## Part 2: Reshaping data


Creating plots and calculating particular statistics often requires data to be in a particular format. Reshaping data is the processing of changing the number of rows and columns of a data table in order to make particular computations or plots easier. The tidyr package in R is package that is useful for reshaping data. Let's explore it now!





$\\$






#### Part 2.1: Simple reshaping


Let's start with the sample example that is in the class slides. The code below recreates this simple data frame. Let's reshape it by: 

1. Converting it into a longer format data frame using the `pivot_longer()` function. 

2. Converting it back to a wider data frame using the `pivot_wider()` function. 


```{r, simple_reshaping}


# the data example from the class slides
the_data <- data.frame(Person = c("Bob", "Alice", "Steve"),
                       Age = c(32, 24, 64), 
                       Height = c(72, 65, 70))


# convert it to a long format


# convert it back to wide format



```




$\\$




#### Part 2.2: Reshaping data to create box plots of book prices


Let's examine the Amazon book price data to practice transforming data using tidyr's `pivot_longer()` and `pivot_wider()` functions. The R chunk below loads the data. 

Let's start by using dplyr to create a simplified data set that only has the columns Title, ISBN.10, Amazon.Price, and List.Price.


```{r}


# download the amazon data
# SDS230::download_data("amazon-books.txt")

# load the book price data and removing NAs
amazon <- read.csv("amazon-books.txt", sep = "\t") %>%
  filter(!is.na(NumPages)) %>%
  filter(!is.na(List.Price)) 


# select only the Title, ISBN.10, Amazon.Price, List.Price variables




```





$\\$





#### Part 2.3: Creating longer data

If we want to create a side-by-side box plot in ggplot we would need a data frame with the following columns:
  1. A column with the categorical variables saying if the price is Amazon's price or the List price
  2. A column with the prices

Let's transform the data to make it longer so that it has these columns using the `pivot_longer(df, cols)` function.


```{r}

# create a long data frame called amazon_long



# let's now create side-by-side boxplots on the long data




```





$\\$






#### Part 2.4: Creating wider data

We can convert data from a long format into a wide format using the tidyr R function `pivot_wider(df, names_from, cols_from)`. 

Let's transform our long data back to wide format. We could imagine a scenario where this would be useful, for example if we wanted the difference in the Amazon vs. List price and our data had come in the initial longer format, then transforming to the wide format would be useful. 


```{r}

# create a long data frame called amazon_long


# let's get the difference in prices and plot these differences in a single boxplot




```



$\\$





## Part 3: Permutation tests for comparing whether more than two means are equal



A group of Hope College students wanted to see if there was an association between a student’s major and the time it takes to complete a small Sudoku-like puzzle

They grouped majors into four categories
 * Applied science (as)
 * Natural science (ns)
 * Social science  (ss)
 * Arts/humanities (ah)





$\\$







##### Step 1: State the null and alternative hypotheses 






$\\$






##### Step 2a: Plot the data


```{r sudoku_plot}


# get the data
sudoku_data <- read.table('http://www.math.hope.edu/isi/data/chap9/MajorPuzzle.txt', header = TRUE)



# create a boxplot of the data using base R




# create a boxplot of the data using ggplot





```




$\\$





##### Step 2b: Calculate the observed statistic




```{r sudoku_obs_stat}


# create a function to get the group range statistic
get_group_range <- function(sudoku_df) {
  

  
  
  
  
}  # end of the function 






# apply this function to our data to get our observed statistic 


  
```






$\\$




##### Step 3: Create the null distribution


```{r sudoku_null_dist, message = FALSE}


# create the null distribution here using a for loop and the get_group_range() function 
# to save time, only create 1,000 points in the null distribution







# plot the null distribution as a histogram





```





$\\$






##### Step 4: Calculate a p-value


```{r sudoku_p_value}



# plot the null distribution again with a line at the observed statistic





# calculate the p-value





```


$\\$







##### Step 5: Make a decision







$\\$








