---
title: "Class 23 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

SDS230::download_data("IPED_salaries_2016.rda")


```





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * One-way ANOVA
 * Connections between ANOVA and linear regression
 * Pairwise comparisons after running an ANOVA




$\\$






## Part 1: One-way ANOVA to analyze if the mean salary differs for different faculty ranks



Let's assess an obvious question: do professors of different ranks have different salaries on average?



$\\$





### Part 1.1: State the null and alternative hypotheses 


Let's start as always by stating the null and alternative hypotheses:

$H_0: \mu_{assistant} = \mu_{associate} = \mu_{full}$

$H_A: \mu_{i} \ne \mu_{j}$ for some pair of i, j


$\alpha = 0.05$




$\\$





### Part 1.2a: Plot the data


```{r IPED_plot}

library(dplyr)

load("IPED_salaries_2016.rda")

IPED_2 <- IPED_salaries |>
  filter(endowment > 0) |>
  mutate(log_salary = log10(salary_tot)) |>
  filter(CARNEGIE %in% c(15, 31)) |>
  filter(rank_name %in% c("Assistant", "Associate", "Full")) |>
  group_by(school) |>
  mutate(num_ranks = n()) |>
  filter(num_ranks == 3)     # only use schools that have all three ranks

dim(IPED_2)



# could look at the log salary instead...
# IPED_2$salary_tot <- IPED_2$log_salary


# create a boxplot of the data using ggplot
IPED_2 |>
  ggplot(aes(rank_name, salary_tot, fill = rank_name)) + 
  geom_boxplot() + 
  xlab("Salary ($)") + 
  ylab("Rank")


```




```{r}

# let's get summary statistics of the data
faculty_summary <- IPED_2 |>
  group_by(rank_name) |>
  summarize(SD_salary = sd(salary_tot), 
            # SD_log_salary = sd(log_salary),   # if we are really worried about heteroscedasticity we could analyze the log of salaries
            salary_tot = mean(salary_tot))


faculty_summary


# let's create another visualization of the data
IPED_2 |>
  ggplot(aes(rank_name, salary_tot, col = rank_name)) + 
  geom_jitter(position = position_jitter(width = .2)) + 
  xlab("Faculty rank") + 
  ylab("Salary ($)") + 
  geom_crossbar(data = faculty_summary, mapping = aes(ymin = salary_tot, ymax = salary_tot, col = rank_name), size=.5, width = .5)


```





$\\$







### Part 1.2b: Calculate the observed statistic

Our observed statistic is an F-statistic:

$$F = \frac{\frac{1}{K-1}\sum_{i=1}^K n_i(\bar{x}_i - \bar{x}_{tot})^2}{\frac{1}{N-K}\sum_{i=1}^K \sum_{j=1}^{n_i} (x_{ij} - \bar{x}_i)^2}$$


We will cheat and use the `lm()` and `anova()` functions to get the F-statistic.
On the homework you will need to calculate this statistic from the data using
dplyr!


```{r IPED_obs_stat}


# Getting the observed F-statistic for the IPED data using built in R functions
# On the homework be sure to use dplyr to actually calculate this statistic!

(obs_stat <- anova(lm(salary_tot ~ rank_name, data = IPED_2))$F[1])


  
```






$\\$






### Step 3: Create the null distribution


Let's visualize the null distribution.


```{r IPED_null_dist, message = FALSE}

# calculate the degrees of freedom
N <- nrow(IPED_2)
K <- 3

(df1 <- K - 1)
(df2 <- N - K)


# visualize the null distribution 
x <- seq(-.5, 5, by = 0.01)
y <- df(x, df1, df2)

plot(x, y, type = "l")



```




$\\$




### Part 1.4: Calculate a p-value


```{r IPED_pvalue}


# calculate the p-value
(p_value <- pf(obs_stat, df1, df2, lower.tail = FALSE))


```





$\\$






### Step 1.5: Make a decision


Since `r p_value` is less than $\alpha = 0.05$ we can reject the null
hypothesis, although we should really check the model assumptions before making
this claim. We will do this next...






$\\$









## Part 2: Connections between ANOVA and regression 


Let's look at connections between the least squares fit we used when fitting
linear regression models and our one-way ANOVA.



$\\$



### Step 2.1: Least squares offsets are the group means


Let's look at the mean salary for each rank and compare it to the
least squares offsets that the `lm()` function finds.


```{r least_squares_means}

# get the mean and sd of the salary for each faculty rank
IPED_stats <- IPED_2 |>
  group_by(rank_name) |>
  summarize(mean_salary = mean(salary_tot),
            sd_salary = sd(salary_tot))


# fit a linear model 
fit_salary <- lm(salary_tot ~ rank_name, data = IPED_2)



# check that the least squares fit offsets are the means of each group
(summary_salary <- summary(fit_salary))

fit_coefs <- coef(fit_salary)


c(fit_coefs[1], fit_coefs[1] + fit_coefs[2], fit_coefs[1] + fit_coefs[3])

IPED_stats$mean_salary

```



$\\$




### Step 2.2: Using R to get an ANOVA table and to look at diagnostic plots


We can use the `anova()` function to create an ANOVA table, and we can use the
`plot()` function to look at diagnostic plots to make sure our ANOVA conditions
have been met.



```{r}

# an easy way to get the ANOVA table using the ANOVA function 
(anova_table <- anova(fit_salary))

# check that SST = SSG + SSE
(SST <- sum((IPED_2$salary_tot - mean(IPED_2$salary_tot))^2))   # SST

sum(anova_table$`Sum Sq`)  #  SSG + SSE



# we can use regression diagnostic plots to assess if ANOVA conditions have been met
par(mfrow = c(2, 2))
plot(fit_salary)


# we should also check that the maximum and minimum standard deviations are not greater 
# than a factor of 2 apart
IPED_stats$sd_salary
max(IPED_stats$sd_salary)/min(IPED_stats$sd_salary)


```




$\\$






###  Part 2.3 Kruskal–Wallis test to see if any of our groups stochastically dominate another group. 

If we are concerned that our one-way ANOVA conditions are not met, we can run a
Kruskal–Wallis test which does not rely on the assumptions of normality and
homoscedasticity. We could also run a permutation test which does not rely on
these assumptions either.


```{r}

#Kruskal–Wallis test
kruskal.test(salary_tot ~ rank_name, data = IPED_2)


# compare to the ANOVA 
# anova(fit_salary)


```





$\\$






## Part 3: Pairwise comparisons after running a one-way ANOVA





###  Part 3.1 Pairwise comparisons in R

If we run a one-way ANOVA and the results are statistically significant, there
are a number of tests we can run to see which pairs of results are significantly
different.



```{r}


# test with no multiple comparisons adjustment (not great)
(pairwise_ttests <- pairwise.t.test(IPED_2$salary_tot, IPED_2$rank_name, p.adj = "none"))


# with the Bonferroni correction
(bonferroni <- pairwise.t.test(IPED_2$salary_tot, IPED_2$rank_name, p.adj = "bonferroni"))


# Note, the Bonferroni p-values are 3 times larger than the p-values with no adjustment
bonferroni$p.value/pairwise_ttests$p.value


# Tukey's HSD test using the TukeyHSD() function 
# It is giving results similar to the Bonferroni correction
fit_salary_aov <- aov(salary_tot ~ rank_name, data = IPED_2)
TukeyHSD(fit_salary_aov)


```






$\\$













