---
title: "Class 8 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)


options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Permutation test for two means using a t-statistic
 * Parametric tests for two means
 * Permutation tests for more than two means (if there is time)





$\\$




## Part 1: Permutation tests for comparing two means in R


Let's us examine the randomized comparative experiment by Lyle et al (1987) to assess whether calcium lowered blood pressure.






##### Step 1: State the null and alternative hypotheses 


$H_0: \mu_{treat} - \mu_{control} = 0$
$H_A: \mu_{treat} - \mu_{control} > 0$


$\alpha = 0.05$




$\\$






##### Step 2a: Plot the data


```{r pill_plot}



treat <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1,  12,  -1,  -3,   3,  -5,   5,   2, -11,  -1,  -3)



# plot the data
boxplot(treat, control, 
        names = c("Treatment", "Control"),
        main = "Comparison of decrease in blood pressure",
        ylab = "Blood pressure decrease")



# create a stripchart
data_list <- list(treat, control) 
  
stripchart(data_list, 
           group.names = c("treatment", "control"), 
           main = "Comparison of decrease in blood pressure",
           xlab = "Blood pressure decrease", 
           col = c("red", "blue"))


```





$\\$





##### Step 2b: Calculate the observed statistic using a t-statistic


The formula for a t-statistic is: 

$t = \frac{\bar{x}_t - \bar{x}_c}{\sqrt{\frac{s^2_t}{n_t} + \frac{s^2_c}{n_c}}}$




```{r pill_obs_stat}






```


$\\$





##### Step 3: Create the null distribution using a permutation/randomization test



```{r null_dist_pill}


# combine the data from the treatment and control groups together


# use a for loop to create shuffled treatment and control groups and shuffled statistics 


       # shuffle data


        # create fake treatment and control groups


        # calculate the t-statistic


        # save the statistic of interest


        



# plot the null distribution as a histogram





```




$\\$





##### Step 4: Calculate a p-value


```{r p_value_pill}


# plot the null distribution again with a red line a the value of the observed statistic




# calculate the p-value





```




$\\$





##### Step 5: Make a decision





When we used a statistic of $\bar{x}_t - \bar{x}_c$ in our randomization test in class 7 we got a p-value of 0.066. How do these results compare? 





$\\$










## Part 2: Parametric tests for comparing two means in R


Let's redo this analysis using a parametric probability distribution, which in this case is the t-distribution. The same 5 steps of hypothesis testing apply here as well!




$\\$




##### Step 1: State the null and alternative hypotheses 


Same as before...


$H_0: \mu_{treat} - \mu_{control} = 0$
$H_A: \mu_{treat} - \mu_{control} > 0$


$\alpha = 0.05$




$\\$





##### Step 2b: Calculate the observed statistic using a t-statistic


Same as before: 


$t = \frac{\bar{x}_t - \bar{x}_c}{\sqrt{\frac{s^2_t}{n_t} + \frac{s^2_c}{n_c}}}$




```{r pill_obs_stat2}


numerator <- mean(treat) - mean(control)

demoninator <- sqrt(   var(treat)/length(treat) + var(control)/length(control)    )


(obs_stat <- numerator/demoninator)


```




$\\$




##### Step 3: Create the null distribution



We will now use a parametric t-distribution (i.e., density function) as a null distribution. The t-distribution has one parameter called "degrees of freedom". We will set this parameter as the minimum of $n_t - 1$ or $n_c - 1$. 


What are the degrees of freedom for this calcium study? 


Let's visualize the t-distribution


```{r t_density}



# plot the density function




# how does this compare to our t-distribution created by shuffling? 



```


$\\$




##### Step 4: Calculate a p-value


We can get $Pr(X > stat)$ for a t-distribution using the `pt()` function.



```{r p_value_pill2}






```





$\\$


##### Step 5: Make a decision


How does our p-value and decision compare to the p-value decision we got from the permutation test? 




$\\$




##### Built in R functions for t-tests

We can use the built in `t.test()` function to run a t-test as well. We will use the extra argument `alternative = "greater"` to do a one-sided t-test. 



```{r t_test}




```


Why is the p-value slightly different than what we got when we used the `pf()` function? 



```{r t_test_sw}



```




See the [Welch–Satterthwaite equation](https://en.wikipedia.org/wiki/Welch%E2%80%93Satterthwaite_equation)




$\\$






## Part 3: Permutation tests for more than two sample means in R




A group of Hope College students wanted to see if there was an association between a student’s major and the time it takes to complete a small Sudoku-like puzzle

They grouped majors into four categories
 * Applied science (as)
 * Natural science (ns)
 * Social science  (ss)
 * Arts/humanities (ah)





$\\$




##### Step 1: State the null and alternative hypotheses 







$\\$






##### Step 2a: Plot the data


```{r sudoku_plot}


# get the data
sudoku_data <- read.table('http://www.math.hope.edu/isi/data/chap9/MajorPuzzle.txt', header = TRUE)



# create a boxplot of the data



```






$\\$







##### Step 2b: Calculate the observed statistic




```{r sudoku_obs_stat}


# we can use the by() function to get the means separately for each group
# (we will learn of a better way to do this later in the semester)
group_means <- as.vector(by(sudoku_data$time, sudoku_data$major, mean))




# a function to get the MAD statistic
get_MAD_stat <- function(group_means) {
  
  total <- 0
  
  for (iGroup1 in 1:(length(group_means) - 1)) {
    for (iGroup2 in (iGroup1 + 1):(length(group_means))){
      total <- total + abs(group_means[iGroup1] - group_means[iGroup2])
    }
  }
  
  total
  
}  # end of the function 




# apply this function to our data to get our observed statistic 

  

```






$\\$







##### Step 3: Create the null distribution


```{r sudoku_null_dist}


# create the null distribution here using a for loop and the get_MAD_stat() function 






# plot the null distribution as a histogram




```




$\\$






##### Step 4: Calculate a p-value


```{r sudoku_p_value}





# plot the null distribution again with a line at the observed statistic





# calculate the p-value






```





$\\$







##### Step 5: Make a decision






$\\$

