---
title: "Class 20 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$




<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
#SDS230::update_installed_packages()

SDS230::download_data("IPED_salaries_2016.rda")



```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Polynomial regression
 * Comparing models with R^2, adjusted R^2, AIC and BIC
 * Cross-validation
 * Visualizing multiple regression models with ggplot



$\\$


## Part 1: Polynomial regression

We can also fit models that have non-linear transformations to our explanatory
variables. This allows us to fit a much broader range of models to our data. In
particular, we can fit models that are polynomial functions of our original
explanatory variables:

$y = \hat{\beta}_0 + \hat{\beta}_1 \cdot x + \hat{\beta}_2 \cdot x^2 + ... + \hat{\beta}_k \cdot + x^k$



$\\$



Let's start by recreating our assistant professor salary data...


```{r}


load("IPED_salaries_2016.rda")


# get the assistant professor data
# we will keep all the variables and all schools in the data frame for this analysis
assistant_data <- filter(IPED_salaries,  endowment > 0,  rank_name == "Assistant")  



```



$\\$




### Part 1.1: Non-linear transformations of the explanatory variable x

Let's explore fitting a model that predicts salary from log endowment, log
endowment squared and log endowment cubed using our salary data from assistant
professors:

$\text{salary} = \hat{\beta}_0 + \hat{\beta}_1 \cdot \text{log(endowment)} + \hat{\beta}_2 \cdot \text{log(endowment)}^2 + \hat{\beta}_2 \cdot + \text{log(endowment)}^3$

To do this we can use the `poly()` function which creates polynomial expansions
of our explanatory variables that are "orthogonal", which means that they will
not have multicolinearity. If we had instead just created polynomial expansions
of our data using say a dplyr `mutate()` operation, then the predictor values
could be highly correlated and thus lead to multicolinearity (e.g., if x = 1:10,
then x and x^2 will be highly correlated leading to multicolinearity).

Let's fit a degree 3 polynomial linear regression model to our data. 


```{r}

# refit our original degree 1 model




# fit a cubic model




# get statistics on our cubic fit



# compare the r^2





```

**Question:** Are the higher order terms statistically significant? 





$\\$





### Part 1.2: Let's visualize our model



```{r}


# create a data frame with a sequence of x-values to make predictions on




# plot the data





# plot the polynomial predictions





```


**Question:** Does this model visually appear to fit the data better? 




$\\$





### Part 1.3: Using an ANOVA to compare the nested models

Because our linear model is nested within our cubic model, we can use an ANOVA
to evaluate whether our cubic model accounts for more of the variability than
our linear model.


```{r}




```



$\\$




### Part 1.4: Higher order polynomials


Let's try fitting even higher order polynomials!


```{r}


# try degrees, 5, 10, etc. 



  # fit a cubic model









```



$\\$




### Part 1.5: Visualizing higher order polynomial fits



Let's visualize our 50 degree polynomial fit



```{r, warning = FALSE}


# fit 50 degree polynomial



# data frame with x-values to get predicted y-hat values 



# predicted y values (i.e., the regression curve)



# visualize the data





```




What would Michael Jackson say about this?  

https://www.youtube.com/watch?v=DQWI1kvmwRg






$\\$








## Part 2:  Comparing models with R^2, adjusted R^2, AIC, BIC and cross-validation


We can compare models using several statistics including $R^2$, $R^2_{adj}$,
$AIC$, $BIC$ and cross-validation. Let's try comparing models using these
statistics to see how well each statistic captures our intuition about the best
model to use.





$\\$







### Part 2.1: Comparing models with R^2, adjusted R^2, AIC, BIC


Let's compare models using these statistics: 

 * `R^2`: where a higher value means a better fit
 * `Adjusted R^2`: where higher value means a better fit
 * `AIC`: where a lower value means a better model
 * `BIC`: where a lower value means a better model

We can use either the synthetic x, y, data we created above, or the assistant
professor data. We will fit the model with different degree polynomials and wee
which modle each statistic selects.


```{r}


library(dplyr)


# fit models of degree 1, 3, and 5





# summarizing these models






# printing the R^2 values (higher is better)
cat('R^2 \n')






# printing the adjusted R^2 values (higher is better)
cat("\n Adjusted R^2 \n")






# printing the AIC values (lower is better)
cat("\n AIC \n")



  

# printing the BIC values (lower is better)
cat("\n BIC \n")






```





$\\$







### Part 2.2: Comparing models using cross-validation


Let's now compare the models using cross-validation. To keep things simple, we
are just going to split the data once into a training set and a test set rather
than do k-fold cross-validation.

We will compare the models based on their mean squared prediction error (MSPE)
where a smaller MSPE is better.



```{r}


# create the training set and the test set






# fit both models on the training data, and calculate the MSPE based on their predictions on the test set









# use a for loop to compare the MSPE for models of degree 1 to 7









```



$\\$



## Part 3: Visualizing multiple regression models with ggplot

In prior classes we have visualized multiple regression models with categorical
predictors using base R graphics. In particular, we created scatter plots for
data in different categories using the `plot()` and `points()` and used the
`col` argument to color the points. We then added on regression lines using the
`abline()` function. This method was useful for educational purposes so that we
could see the connection between the model parameters that were estimated using
the `lm()` function, and the underlying data. However, if we want to create
better looking visualizations in a more efficient manner, then it is better to
use ggplot.

Let's now use ggplot to visualize a multiple regression model that has one
quantitative and one categorical variable. In particular, let's recreate the
plot from class 19 part 3.6 where we display faculty salaries as a function of
log endowment separately for different faculty ranks.



```{r}

# load the data into R
load("IPED_salaries_2016.rda")


# create the IPED data subset used in class 19 part 3
IPED_2 <- IPED_salaries |>
  filter(endowment > 0) |>
  mutate(log_endowment = log10(endowment)) |>
  filter(CARNEGIE %in% c(15, 31)) |>
  filter(rank_name %in% c("Assistant", "Associate", "Full"))

dim(IPED_2)



# using ggplot






```



$\\$








