---
title: "Class 8 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)


options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Permutation test for two means using a t-statistic
 * Parametric tests for two means




$\\$




## Part 1: Permutation tests for comparing two means in R


Let's us reexamine the randomized comparative experiment by Lyle et al (1987) to assess whether calcium lowered blood pressure.






### Step 1: State the null and alternative hypotheses 


$H_0: \mu_{treat} - \mu_{control} = 0$
$H_A: \mu_{treat} - \mu_{control} > 0$


$\alpha = 0.05$




$\\$






### Step 2a: Plot the data


```{r pill_plot}



treat <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1,  12,  -1,  -3,   3,  -5,   5,   2, -11,  -1,  -3)



# plot the data
boxplot(treat, control, 
        names = c("Treatment", "Control"),
        main = "Comparison of decrease in blood pressure",
        ylab = "Blood pressure decrease")



# create a stripchart
data_list <- list(treat, control) 
  
stripchart(data_list, 
           group.names = c("treatment", "control"), 
           main = "Comparison of decrease in blood pressure",
           xlab = "Blood pressure decrease", 
           col = c("red", "blue"))


```





$\\$





### Step 2b: Calculate the observed statistic using a t-statistic


The formula for a t-statistic is: 

$t = \frac{\bar{x}_t - \bar{x}_c}{\sqrt{\frac{s^2_t}{n_t} + \frac{s^2_c}{n_c}}}$




```{r pill_obs_stat}


numerator <- mean(treat) - mean(control)

demoninator <- sqrt(   var(treat)/length(treat) + var(control)/length(control)    )


(obs_stat <- numerator/demoninator)


```


$\\$





### Step 3: Create the null distribution using a permutation/randomization test



```{r null_dist_pill}


# combine the data from the treatment and control groups together
combined_data <- c(treat, control)


# use a for loop to create shuffled treatment and control groups and shuffled statistics 
null_distribution <-  NULL
for (i in 1:10000) {

       # shuffle data
        shuff_data <- sample(combined_data)

        # create fake treatment and control groups
        shuff_treat   <-  shuff_data[1:10]
        shuff_control  <-  shuff_data[11:21]

        numerator_shuff <- mean(shuff_treat) - mean(shuff_control)
        demoninator_shuff <- sqrt(var(shuff_treat)/length(shuff_treat) + 
                                  var(shuff_control)/length(shuff_control))

        # save the statistic of interest
        null_distribution[i] <- numerator_shuff/demoninator_shuff

        
}


# plot the null distribution as a histogram
hist(null_distribution, 
     breaks = 100,
     main = "Null distribution", 
     xlab = "t-statistic")




```



$\\$





### Step 4: Calculate a p-value


```{r p_value_pill}


# plot the null distribution again with a red line a the value of the observed statistic
hist(null_distribution, 
     breaks = 100,
     main = "Null distribution", 
     xlab = "t-statistic")


abline(v = obs_stat, col = "red")



# calculate the p-value
(p_value <- sum(null_distribution >= obs_stat)/10000)




```



$\\$



### Step 5: Make a decision

Since `r p_value` is greater than $\alpha = 0.05$ we can not reject the null hypothesis. Thus if we are using the Neyman-Pearson paradigm, we do not have sufficient evidence to say that the pill is effective. 



When we used a statistic of $\bar{x}_t - \bar{x}_c$ in our randomization test in class 7 we got a p-value of 0.066. How do these results compare? 





$\\$










## Part 2: Parametric tests for comparing two means in R


Let's redo this analysis using a parametric probability distribution, which in this case is the t-distribution. The same 5 steps of hypothesis testing apply here as well!




$\\$




### Step 1: State the null and alternative hypotheses 


Same as before...


$H_0: \mu_{treat} - \mu_{control} = 0$
$H_A: \mu_{treat} - \mu_{control} > 0$


$\alpha = 0.05$




$\\$





### Step 2b: Calculate the observed statistic using a t-statistic


Same as before: 


$t = \frac{\bar{x}_t - \bar{x}_c}{\sqrt{\frac{s^2_t}{n_t} + \frac{s^2_c}{n_c}}}$




```{r pill_obs_stat2}


numerator <- mean(treat) - mean(control)

demoninator <- sqrt(   var(treat)/length(treat) + var(control)/length(control)    )


(obs_stat <- numerator/demoninator)


```




$\\$




### Step 3: Create the null distribution



We will now use a parametric t-distribution (i.e., density function) as a null distribution. The t-distribution has one parameter called "degrees of freedom". We will set this parameter as the minimum of $n_t - 1$ or $n_c - 1$. 


What are the degrees of freedom for this calcium study? 


Let's visualize the t-distribution


```{r t_density}


(degree_free <- min(length(treat), length(control)) - 1)

x_vals <- seq(-10, 10, by = .01)
y_vals <- dt(x_vals, degree_free)

plot(x_vals, y_vals, type = "l")


# how does this compare to our t-distribution created by shuffling? 

hist(null_distribution, freq = FALSE,
     breaks = 100,
     main = "Null distribution", 
     xlab = "t-statistic")

points(x_vals, y_vals, type = "l", col = "red")


```


$\\$




### Step 4: Calculate a p-value


We can get $Pr(X > stat)$ for a t-distribution using the `pt()` function.



```{r p_value_pill2}


(p_val <- pt(obs_stat, degree_free, lower.tail = FALSE))


```





$\\$


### Step 5: Make a decision


How does our p-value and decision compare to the p-value decision we got from the permutation test? 




$\\$




### Built in R functions for t-tests

We can use the built in `t.test()` function to run a t-test as well. We will use the extra argument `alternative = "greater"` to do a one-tailed t-test. 



```{r t_test}

t.test(treat, control, alternative = "greater")


```


Why is the p-value slightly different than what we got when we used the `pt()` function? 








