---
title: "Class 5 notes and code"
output:
  pdf_document: default
  html_document: default
---



$\\$




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)

knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```





<!--  Please run the code in the R chunk below once. It will install a packages, and and download figures/data we will use.  -->

```{r, eval = FALSE, echo = FALSE}


# download the okcupid profile data again
download.file("https://raw.githubusercontent.com/emeyers/SDS230/master/ClassMaterial/data/profiles_revised.csv", "profiles_revised.csv", mode = "wb")


```





$\\$




## Overview

 * Simulating sampling distributions 
 * Confidence intervals
 * The bootstrap
 * Using the bootstrap to calculate confidence intervals in R



$\\$








## Part 1: Sampling distributions and the standard error



#### Part 1.1: Review:  Sampling distributions


A distribution of statistics is called a **sampling distribution**. 

Let's use simulations to generate an approximation to a sampling distribution (a true sampling distribution would have all possible statistic values in it). 


```{r}


# create a sampling distribution of the mean using data from a uniform distribution

sampling_dist <- NULL







# plot a histogram of the sampling distribution of these means 






```


For many statistics, and the sample average $\bar{x}$ in particular, our sampling distribution is normal. 

We will discuss ways to visualize whether data comes from a particular distribution later in the semester. 



$\\$





#### Part 1.2: The standard error (SE)

The deviation of a sampling distribution is called the standard error (SE). Can you calculate the standard error for the sampling distribution you created above? 

```{r}

# caclulate the SE as the standard deviation of the sampling distribution




```






$\\$






### Part 1.3: Approximating a sampling distributions from a data set


We generate samples from an actual data set we have using the `sample()` function.  

Let's start by just generate a single sample of size n = 100 from the OkCupid users' heights and calculating the mean of this sample.


```{r okcupid_sampling_dist1}

# read in the okcupid data
profiles <- read.csv("profiles_revised.csv")


# get the heights for the OkCupid data



# get one random sample of heights from 100 people



# get the mean of this sample



```





$\\$





We can then create an approximation of a sampling distribution from the OkCupid users' data set by repeating this many times in a for loop.



```{r okcupid_sampling_dist2}


# repeat the process 1,000 times
sampling_dist <- NULL 






# plot a histogram of this sampling distribution 




```




**Question:** What would have to be true for this to be an actual sampling distribution? 







$\\$









## Part 2:  The central limit theorm

The [central limit theorm (CTL)](https://en.wikipedia.org/wiki/Central_limit_theorem) establishes that (in most situations) when independent random variables are added their (normalized) **sum** converges to a normal distribution.

Put another way, if we definte the average random (i.i.d) sample {$X_1$, $X_2$, ..., $X_n$} of size *n* as: 

$S_{n}:={\frac{X_{1}+\cdots +X_{n}}{n}}$

then the CTL tells us that:

$\sqrt{n}(S_{n} - \mu)$ $\xrightarrow {d} N(0,\sigma^{2})$


You will explore this more through simulations on homework 2. 






$\\$








## Part 3: Confidence intervals


$\\$


Confidence intervals are ranges of values that capture a parameter a fix proportion of time.




$\\$






## Part 4: The bootstrap

The bootstrap is a method that can be used to create confidence intervals for a large range of parameters. 

The central concept behind the bootstrap is the "plug-in principle" where we treat our sample of data as if it were the population. We then sample *with replacement* from **our sample** to create a *bootstrap distribution* which is a proxy for (the spread of) the sampling distribution. 




$\\$




#### Part 4.1: Creating a bootstrap distribution in R


To sample data in R we can use the `sample(the_data)` function. To sample data with replacement we use the `replace = TRUE` argument, i.e., `sample(the_data, replace = TRUE)`.  

Below we calculate the bootstrap distribution for mean age of OkCupid users using just the first 20 OkCupid users in the data set. 


```{r boot_dist}


# read in the okcupid data
profiles <- read.csv("profiles_revised.csv")



# get the ages from the first 20 OkCupid profiles





# create the bootstrap distribution






# plot the bootstrap distribution to make sure it looks normal





```



$\\$




#### Part 4.2: Calculating the bootstrap standard error SE*


The standard deviation of the *bootstrap distribution* is usually a good approximation of the standard deviation of the sampling distribution - i.e, it is a good approximation of the *standard error SE*.

When our bootstrap distribution is relatively normal, we can use the fact that 95% of values fall within to standard deviations of a normal distribution to calculate 95% confidence intervals as:

$CI_{95} = [stat - 2 \cdot SE^*,  stat + 2 \cdot SE^*]$

For example, for a our bootstrap distribution we have a 95% confidence interval for the mean $\mu$ as: 

$CI_{95} = [\bar{x} - 2 \cdot SE^*,  \bar{x} + 2 \cdot SE^*]$




```{r bootstrap_SE}

# calculate the bootstrap standard error SE* as the standard deviation of the bootstrap distribution 



# calculate the 95% CI using SE*





```



Above we are using the bootstrap to create a 95% confidence interval which should capture the mean age $\mu_{age}$ 95% of the time.









