---
title: "Class 13 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$







<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


SDS230::download_data("state_demographics.rda")


```





```{r setup, include=FALSE}


library(latex2exp)
library(ggplot2)
library(dplyr)

options(scipen=999)
knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```





$\\$







## Overview

 * Mapping
 * Simple linear regression
 * ggplot bonus features






$\\$







## Part 1: Creating choropleth maps



To practice creating choropleth maps, we will examine demographic patterns in the United States by creating a map of people who are of retirement age. 


For more fun data sets to map, check out [howmuch.net](https://howmuch.net/sources/income-inequality-by-state)





$\\$






#### Part 1.1: Get map coordinates



To start our mapping practice, let's plot Yale's location (longitude = -72.9223, latitude = 41.3163) on a map of Connecticut.



```{r}


# Get the map of the States
states_map <- map_data("state") 
  

# Plot a state
states_map %>%
  filter(region == "connecticut") %>%
  ggplot(aes(x = long, y = lat)) + 
  geom_polygon() + 
  geom_point(aes(x = -72.9223, y = 41.3163), col = "red")   # add Yale


```





$\\$






#### Part 1.2: Combine demographic data with the map coordinates


Let's now look at state demographic information and combine this with mapping data.



```{r}


# load the data
load("state_demographics.rda")


# mutate the data to names of states in lower case
state_demographics <- state_demographics %>%
  mutate(state_lower = tolower(State))


# join state_demographics and states_map data frames 
states_map <- map_data("state") %>%
  left_join(state_demographics, by = c("region" = "state_lower"))


# order the data
states_map <- arrange(states_map, group, order)

```





$\\$





#### Part 1.3: Plot the map


To get a sense of which states have an older population, let's plot the number of people in each state who are over 64 years old. 



```{r}

ggplot(states_map, aes(x = long, y = lat, group = group, fill = over_64)) + 
      geom_polygon(color = "black") +   
      coord_map("polyconic")    # plots it on a curved earth map


```



**Question:**  What is wrong with this figure? 





$\\$






#### Part 1.4: Normalizing data


In order to really see which states have older populations, we need to normalize our maps based on the total population. 

Do the results from the normalized map make more sense? 


```{r}


states_map <- states_map %>%
  mutate(prop_over_64 = over_64/total)

ggplot(states_map,  aes(x = long, y = lat, group = group,  fill = prop_over_64)) +   
	geom_polygon(color = "black") + 
	coord_map("polyconic") +   
  scale_fill_gradient(low = "black", high = "red")



```






$\\$







## Part 2: Simple linear regression in R


As you probably are aware, the 2020 presidential election is coming up on November 3rd. Many people are anxious to know the outcome of the election. Let's build an simple linear regression model that can predict the margin a president will win (or lose) an election by as a function of their approval rating on election day. 

To build the this model, we will use data from the previous 12 elections in which an incumbent was running for reelection. The data is from this [FiveThirtyEight](https://fivethirtyeight.com/features/approval-ratings-and-re-election-odds/) and was compiled for an introductory text book by Lock, et al. 




$\\$





#### Part 2.1: Plotting the data and fitting a simple linear regression model 


Let's start by plotting the data, fitting our linear model, and visualizing the fit!


```{r}


# load the data from the past 12 elections
load(url("https://www.lock5stat.com/datasets/ElectionMargin.rda"))



# create a scatter plot and calculate the correlation (note: this is plot(x, y))
plot(ElectionMargin$Approval, ElectionMargin$Margin,
     xlab = "President approval rating on election day",
     ylab = "President margin of the vote") 


# calculate the correlation
cor(ElectionMargin$Approval, ElectionMargin$Margin)


# fit a regression model  (note: this is y as as function of x)
(lm_fit <- lm(Margin ~ Approval, data = ElectionMargin))


# add the regression line to the plot
abline(lm_fit)

# do any points appear to be outliers? 



# create a scatter plot using ggplot
ElectionMargin %>%
  ggplot(aes(Approval, Margin)) + 
  geom_point() + 
  xlab("President approval rating on election day") + 
  ylab ("President margin of the vote") + 
  geom_smooth(method = "lm", se = FALSE)


```






$\\$




#### Part 2.2: Regression coefficients and residuals



Now let's examine:

1. The regression coefficients found by our model.
2. The predicted values y-hat values for each x value in our data set.
3. The residuals.



```{r}


# examine the beta0 and beta1 coefficients
coef(lm_fit)


# could you write out this linear regression equation? 



# let's look at all the fitted values y-hat for each x in our data set
lm_fit$fitted.values


# let's look at the residuals (y - y-hat)
lm_fit$residuals


# do these residual match what we would expect based on y - fitted.values? 
max((ElectionMargin$Margin - lm_fit$fitted.values) - lm_fit$residuals)



```





$\\$




#### Part 2.3: Making predictions from the model


We can also make predictions $\hat{y}$ for new x-values. Let's predict what Donald Trump's margin of the vote will be based on his current approval rating, which is 43.5% as of 10/6/2020 [according to FiveThirtyEight](https://projects.fivethirtyeight.com/trump-approval-ratings/).



```{r}


# create a new data frame with values we want to predict
trump_approval_df <- data.frame(Approval  = c(43.5, 57))


# predicted y-hat values
(predicted_margin <- predict(lm_fit, newdata = trump_approval_df))


# replot data with these new points on it
plot(ElectionMargin$Approval, ElectionMargin$Margin,
     xlab = "President approval rating on election day",
     ylab = "President margin of the vote") 


# add the regression line to the plot
abline(lm_fit)


points(trump_approval_df$Approval, predicted_margin, col = "red", pch = 19)




```



$\\$





## Part 3: ggplot bonus features



ggplot bonus features:

 * plotly
 * emojis
 * animations




$\\$



#### Part 3.1: Plotly 


The plotly package can be used to create interactive visualizations. 

We can use plotly in conjunction with ggplot by saving our ggplot figure to an object `g`, and then we can use `ggplotly(g)` to create an interactive visualization. 

Let's create an interactive plot of the Margin of the vote presidents' received as a function of their approval rating.


```{r}

library(plotly)

load(url("https://www.lock5stat.com/datasets/ElectionMargin.rda"))


# create an interactive scatter plot using ggplot and plotly
g <- ElectionMargin %>%
  ggplot(aes(Approval, Margin, col = Result)) + 
  geom_point(aes(name = Candidate)) + 
  xlab("President approval rating on election day") + 
  ylab ("President margin of the vote") + 
  geom_smooth(method = "lm", se = FALSE)

#ggplotly(g)


```


$\\$


If you would like additional practice, use the gapminder data to create a plotly scatter plot of data from 2007 using:

 * x = life expectancy
 * y = gpd per capita
 * col = continent
 * name = country

```{r, plotly}

library(gapminder)


g <- gapminder %>% 
  filter(year == 2007) %>%  	
  ggplot(aes(x = gdpPercap, y = lifeExp, col = continent, name = country)) + 	  
	geom_point()

#ggplotly(g)


```





$\\$






#### Part 3.2: emojis



```{r, emoji}

#install.packages("emojifont")

library(emojifont)

load.emojifont('OpenSansEmoji.ttf')


ggplot(mtcars, aes(x = wt, y = mpg, label = emoji("car"), col = factor(cyl))) +  	  	
	geom_text(family="OpenSansEmoji", size=6)  +    
	xlab("Weigth") +   
	ylab("Miles per Gallon") + 
	theme_classic()


```






$\\$






#### Part 3.3: animations


```{r, animations}


#install.packages('png')
#install.packages('gganimate')


# library(gganimate)
# 
# 
# ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
#   geom_point(alpha = 0.7, show.legend = FALSE) +
#   scale_colour_manual(values = country_colors) +
#   scale_size(range = c(2, 12)) +
#   scale_x_log10() +
#   facet_wrap(~continent) +
#   # Here comes the gganimate specific bits
#   labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
#   transition_time(year) +
#   ease_aes('linear')



```




