---
title: "Class 11 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


SDS230::download_data("x_y_join.rda")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)

library(dplyr)


#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```






$\\$






## Overview


 * Plotting using grammar of graphics with ggplot
 * Joining data frames






$\\$








## Part 1: Visualizations with ggplot


We can use the ggplot2 package which is part of the tidyverse to create much nicer looking graphics than using base R graphics. The ggplot2 library is modeled on Leland Wilkinson's "grammar of graphics" which tries to create graphics as a combination of basic visual elements. 


In the exercises below, we will use the motor trends cars data set (mtcars) that comes with base R installation, and also the gapminder data. 





$\\$





#### Part 1.1: Scatter plots


```{r}


# install.packages("ggplot2")


library(ggplot2)


# base R
plot(mtcars$wt, mtcars$mpg)


# ggplot - global mapping
ggplot(data = mtcars, mapping = aes(x = wt, y = mpg)) +
	geom_point() 


# ggplot - shorter global mapping
ggplot(mtcars, aes(x = wt, y = mpg)) +
	geom_point() 


# ggplot - mapping in the geom
ggplot(data = mtcars) +   
	geom_point(mapping = aes(x = wt, y = mpg))


```




$\\$





#### Part 1.2: Adding labels to plots


We can add labels to the plots using the xlab("label1") and ylab("label2") functions



```{r}


ggplot(mtcars, aes(x = wt, y = mpg)) +  	  	
	geom_point()  +    
	xlab("Weight") +   
	ylab("Miles per Gallon")



```



Remember, if you don't want exes [label your axes](https://xkcd.com/833/)




$\\$





#### Part 1.3: More aesthetic mappings


```{r}


# add color based on the transmission type (is automatic or not)
ggplot(mtcars, aes(x = wt, y = mpg, col = am)) +
	geom_point()


# it is better to treat the transmission type as a categorical variable?
ggplot(mtcars, aes(x = wt, y = mpg, col = factor(am))) +  
  geom_point()


# can also try mapping transmission type to shape or size 
ggplot(mtcars, aes(x = wt, y = mpg, shape = factor(am))) +  
   geom_point()
 
ggplot(mtcars, aes(x = wt, y = mpg, size = am)) +  
   geom_point()


# Survey question 4: When adding the variable automatic/manual transmission (am) to the scatter plot mpg vs. weight, do you think it is best to map am on to:  a) color b) shape c) size


```



Survey question 4: When adding the variable automatic/manual transmission (am) to the scatter plot mpg vs. weight, do you think it is best to map am on to: 

a) color
b) shape
c) size




$\\$





#### Part 1.4: Attributes vs. Aesthetics


Setting an **aesthetic mapping** maps a variable to a **glyph property**

Setting an **attribute** set a glyph property to a **fixed value**



```{r}


# setting a aesthetic
ggplot(mtcars) +
	geom_point(aes(x = wt, y = mpg, col = factor(am)))


# setting an attribute
ggplot(mtcars) + 
	geom_point(aes(x = wt, y = mpg), col = "red")


```




$\\$





#### Part 1.5: Facets


Beyond comparing variables based on aesthetics you can compare categorical variables by splitting a plot into subplots (called facets) using facet_wrap()




```{r}


# separate subplots for the two transmission types
ggplot(mtcars, aes(x = wt, y = mpg)) +  	
  geom_point() +   
	facet_wrap(~am)


# One can also do facets in two dimensions
ggplot(mtcars, aes(x = wt, y = mpg)) +  	
	geom_point() +   
	facet_wrap(am ~ cyl)



```




$\\$





#### Part 1.6: Overplotting


Sometimes points overlap making it hard to estimate the number of points at a particular range of values. 

We can control the transparency of points by changing their alpha values.




```{r}

library(gapminder)


# a lot of overplotting
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 	
	geom_point()


# changing the transparency levels makes it a bit easir to see how many points are at a given x, y location
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 	
	geom_point(alpha = .1)



```





$\\$





#### Part 1.7: Changing scales


We can transform the scales of continuous data too.


```{r}

# changing the scale on the x-axis
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 	
	geom_point(alpha = .2) + 
  scale_x_continuous(trans='log10')


```






$\\$






#### Part 1.8: geoms


We can use different geoms to create other types of plots.



```{r}


# Let's create a plot that shows the GDP in the United States as a function of the year using the geom geom_line()
gapminder %>% 
  filter(country == 'United States') %>%  	
  ggplot(aes(x = year, y = gdpPercap)) + 	  
	geom_line()



# Let's plot a histogram of the weights of cars
ggplot(mtcars, aes(x = wt)) + 	  	
	geom_histogram(bins = 10)


# Let's create a boxplot of the weights of cars
ggplot(mtcars, aes(x = "", y = wt)) + 	  	
	geom_boxplot()


# Let's create a side-by-side boxplot of the weights of cars depending on the number of cylinders the engine has
ggplot(mtcars, aes(x = factor(cyl), y = wt)) + 	  	
  geom_boxplot()



```





$\\$





#### Part 1.9: geoms continued

Violin and Joy plots are other ways to view distributions of data



```{r}

# violin plot 
ggplot(mtcars, aes(x = factor(cyl), y = wt)) + 	
	geom_violin()



library("ggridges")

# joy plot
ggplot(mtcars, aes(y = factor(cyl), x = wt)) + 	
  geom_density_ridges()



```


Survey question 5: Can you figure out where the name "joy plot" comes from? 




$\\$





#### Part 1.10: geoms continued


We can also have multiple geom layers on a single graph by using the + symbol
E.g   ggplot(…) + geom_type1() + geom_type2()



```{r}

# Create a scatter plot of miles per gallon as a function of weight and then add a smoothed line using geom_smooth()

ggplot(mtcars, aes(x = wt, y = mpg)) +  	
	geom_point() +  
	geom_smooth() 


```





$\\$





#### Part 1.10: Themes


We can also use different themes to change the appearance of our plot.


```{r}


# Add theme_classic() to our plot

ggplot(mtcars, aes(x = wt, y = mpg)) +  	  	
	geom_point()  +    
	xlab("Weigth") +   
	ylab("Miles per Gallon") + 
	theme_classic()


```




$\\$




#### Part 1.11: Exeriment yourself


Try to create some interesting visualizations from either a data set we have used in the class, or a new data set you found.




```{r}

library(latex2exp)










```










$\\$







## Part 2: Joining data frames


Often data of interest is spread across multiple data frames that need to be joined together into a single data frame for further analyses. We will explore how to do this using dplyr. 


Let's look at a very simple data set to explore joining data frames. 


```{r}

library(dplyr)


load('x_y_join.rda')

x
y



```


$\\$



#### Part 2.1: Left join

Left joins keep all rows in the left table.  

Data from right table added when there is the key matches, otherwise NA as added. 

Try to do a left join of the data frames x and y using their keys.


```{r}

left_join(x, y, by = c("key_x" = "key_y"))


```





$\\$





#### Part 2.2: Right join


Right joins keep all rows in the right table.  

Data from left table added when there is the key matches, otherwise NA as added. 

Try to do a right join of the data frames x and y using their keys.


```{r}

right_join(x, y, by = c("key_x" = "key_y"))


```



$\\$



#### Part 2.3: Inner join


Inner joins only keep rows in which there are matches between the keys in both tables

Try to do an inner join of the data frames x and y using their keys.


```{r}


inner_join(x, y, by = c("key_x" = "key_y"))


```



$\\$



#### Part 2.4: Full join

Full joins keep all rows in both table.  

NAs are added where there are no matches.




```{r}

full_join(x, y, by = c("key_x" = "key_y"))


```



$\\$




#### Part 2.5a: Duplicate keys


Duplicate keys are useful if there is a one-to-many relationship (duplicates are usually in the left table). 

Let's look at two other tables that have duplicate keys



```{r}

x2
y2

nrow(x2)
nrow(y2)


```




$\\$





#### Part 2.5b: Duplicate keys


If both tables have duplicate keys you get all possible combinations (Cartesian product). This is almost always an error! Always check the output dimension after you join a table because even if there is not a syntax error you might not get the table you are expecting!


Try doing a left join on the data frames x2 and y2 using only their first keys (i.e., key1_x and key1_y). Save the joined data frame to an object called `x2_joined`. Note that `x2_joined` has more rows than the original `x2` data frame despite the fact that you did a left join!  This is due to duplicate keys in both x2 and y2. 

Usually a mistake was made when a data frame ends up having more rows after a left join. It is good to check how many rows a data frame has before and after a join to catch any possible errors. 



```{r}

# initial left data frame only has 3 rows
nrow(x2)


# left join when both the left and right tables have duplicate keys
(x2_joined <- left_join(x2, y2, by = c("key1_x" = "key1_y")))


# output now has more rows than the initial table
nrow(x2_joined)


```




$\\$




#### Part 2.5c: Duplicate keys



To deal with duplicate keys in both tables, we can join the tables **using multiple keys** in order to make sure that each row is uniquely specified.

Try doing a left join on the data frames x2 and y2 using both the keys. Save the joined data frame to an object called `x2_joined_mult_keys`. Note that `x2_joined_mult_keys` has the same number of rows as the original `x2` data frame which is usually what we want when we do a left join.




```{r}


# initial left data frame only has 3 rows
nrow(x2)


# join the data frame using multiple keys
x2_joined_mult_keys <- left_join(x2, y2, c("key1_x" = "key1_y", "key2_x" = "key2_y"))

# output now only has 3 rows
nrow(x2_joined_mult_keys)


```



$\\$ 






