---
title: "Class 15 notes and code"
output:
  pdf_document: 
   fig_width: 7
   fig_height: 5
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


SDS230::download_data("amazon-books.txt")


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(tidyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```






$\\$






## Overview

 * Reshaping data
 * Hypothesis tests for more than 2 means





$\\$







## Part 1:  Reshaping data




#### Part 1.1: Simple reshaping


Let's start with the sample example that is in the class slides. The code below recreates this simple data frame. Let's reshape it by: 

1. Converting it into a longer format data frame using the `pivot_longer()` function. 

2. Converting it back to a wider data frame using the `pivot_wider()` function. 


```{r, simple_reshaping}


# the data example from the class slides
the_data <- data.frame(Person = c("Bob", "Alice", "Steve"),
                       Age = c(32, 24, 64), 
                       Height = c(72, 65, 70))


# convert it to a long format


# convert it back to wide format



```




$\\$




#### Part 1.2: Reshaping data to create box plots of book prices


Let's examine the Amazon book price data to practice transforming data using tidyr's `pivot_longer()` and `pivot_wider()` functions. The R chunk below loads the data. 

Let's start by using dplyr to create a simplified data set that only has the columns Title, ISBN.10, Amazon.Price, and List.Price.


```{r}


# download the amazon data
# SDS230::download_data("amazon-books.txt")

# load the book price data and removing NAs
amazon <- read.csv("amazon-books.txt", sep = "\t") %>%
  filter(!is.na(NumPages)) %>%
  filter(!is.na(List.Price)) 


# select only the Title, ISBN.10, Amazon.Price, List.Price variables




```





$\\$





#### Part 1.3: Creating longer data

If we want to create a side-by-side box plot in ggplot we would need a data frame with the following columns:
  1. A column with the categorical variables saying if the price is Amazon's price or the List price
  2. A column with the prices

Let's transform the data to make it longer so that it has these columns using the `pivot_longer(df, cols)` function.


```{r}

# create a long data frame called amazon_long



# let's now create side-by-side boxplots on the long data




```





$\\$






#### Part 1.4: Creating wider data

We can convert data from a long format into a wide format using the tidyr R function `pivot_wider(df, names_from, cols_from)`. 

Let's transform our long data back to wide format. We could imagine a scenario where this would be useful, for example if we wanted the difference in the Amazon vs. List price and our data had come in the initial longer format, then transforming to the wide format would be useful. 


```{r}

# create a long data frame called amazon_long


# let's get the difference in prices and plot these differences in a single boxplot




```



$\\$







## Part 2: Permutation tests for comparing whether more than two means are equal



A group of Hope College students wanted to see if there was an association between a student’s major and the time it takes to complete a small Sudoku-like puzzle

They grouped majors into four categories
 * Applied science (as)
 * Natural science (ns)
 * Social science  (ss)
 * Arts/humanities (ah)





$\\$







##### Step 1: State the null and alternative hypotheses 






$\\$






##### Step 2a: Plot the data


```{r sudoku_plot}


# get the data
sudoku_data <- read.table('http://www.math.hope.edu/isi/data/chap9/MajorPuzzle.txt', header = TRUE)



# create a boxplot of the data using base R




# create a boxplot of the data using ggplot





```




$\\$






##### Step 2b: Calculate the observed statistic



We will run our permutation test using the group range statistic:

$G = \max{\bar{x}_i} - \min{\bar{x}_i}$



As an alternative, we could use an F-statistic (although we won't for now): 

$F = \frac{\frac{1}{K-1}\sum_{i=1}^K n_i(\bar{x}_i - \bar{x}_{tot})^2}{\frac{1}{N-K}\sum_{i=1}^K \sum_{j=1}^{n_i} (x_{ij} - \bar{x}_i)^2}$




```{r sudoku_obs_stat}



# If we wanted to use an F statistic, here is a function that would calculate it
#get_F_stat <- function(sudoku_df) {
#  anova(lm(time ~ major, data = sudoku_df))$F[1]
#}




# create a function to get the group range statistic
get_group_range <- function(sudoku_df) {
  

  
  
  
  
}  # end of the function 






# apply this function to our data to get our observed statistic 


  
```






$\\$




##### Step 3: Create the null distribution


```{r sudoku_null_dist, message = FALSE}


# create the null distribution here using a for loop and the get_group_range() function 
# to save time, only create 1,000 points in the null distribution







# plot the null distribution as a histogram





```





$\\$






##### Step 4: Calculate a p-value


```{r sudoku_p_value}



# plot the null distribution again with a line at the observed statistic





# calculate the p-value





```


$\\$







##### Step 5: Make a decision







$\\$








