---
title: "Class 23 bonus notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()


```




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)
library(dplyr)
library(ggplot2)

#options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * String manipulation



$\\$






## Part 1: String manipulation using stringr


$\\$


##### Part 1.1: Base R and stringr


```{r}

library(stringr)


```



$\\$


##### Part 1.2: Trimming and padding a string


```{r}





```



$\\$



##### Part 1.3: Extracting a piece of a string


```{r}





```



$\\$



##### Part 1.4: Concatenating strings


```{r}





```



$\\$





##### Part 1.5: Download an article



```{r}



base_name <- "https://www.nytimes.com/2020/11/16/us/politics/"
article_name <- "trump-iran-nuclear.html"

full_name <- str_c(base_name, article_name)

#download.file(full_name, article_name)


# viewer <- getOption("viewer")
# viewer(article_name)




```



$\\$


##### Part 1.6: string length



```{r}






```



$\\$



##### Part 1.7: string replace all


```{r}



```




$\\$


##### Part 1.8: String split


```{r}






```




$\\$


##### Part 1.9: Extracting strings


```{r}




```



$\\$


##### Part 1.10: Detecting a string


```{r}



```



$\\$




## Part 2: Regular expressions


We can use regular expressions to do a lot more complex string matching!





