---
title: "Class 5 notes and code"
output:
  pdf_document: default
  html_document: default
---



$\\$




```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)

knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```





<!--  Please run the code in the R chunk below once. It will install a packages, and and download figures/data we will use.  -->

```{r, eval = FALSE, echo = FALSE}


# download the okcupid profile data again
download.file("https://raw.githubusercontent.com/emeyers/SDS230/master/ClassMaterial/data/profiles_revised.csv", "profiles_revised.csv", mode = "wb")


```





$\\$




## Overview

 * The central limit theorem 
 * Confidence intervals
 * The bootstrap
 * Using the bootstrap to calculate confidence intervals in R



$\\$









## Part 0:  The central limit theorm

The [central limit theorm (CTL)](https://en.wikipedia.org/wiki/Central_limit_theorem) establishes that (in most situations) when independent random variables are added their (normalized) **sum** converges to a normal distribution.

Put another way, if we definte the average random (i.i.d) sample {$X_1$, $X_2$, ..., $X_n$} of size *n* as: 

$S_{n}:={\frac{X_{1}+\cdots +X_{n}}{n}}$

then the CTL tells us that:

$\sqrt{n}(S_{n} - \mu)$ $\xrightarrow {d} N(0,\sigma^{2})$


You will explore this more through simulations on homework 2. 






$\\$








## Part 1: Confidence intervals


$\\$


Confidence intervals are ranges of values that capture a parameter a fix proportion of time.




$\\$






## Part 2: The bootstrap

The bootstrap is a method that can be used to create confidence intervals for a large range of parameters. 

The central concept behind the bootstrap is the "plug-in principle" where we treat our sample of data as if it were the population. We then sample *with replacement* from **our sample** to create a *bootstrap distribution* which is a proxy for (the spread of) the sampling distribution. 




$\\$




#### Part 2.1: Creating a bootstrap distribution in R


To sample data in R we can use the `sample(the_data)` function. To sample data with replacement we use the `replace = TRUE` argument, i.e., `sample(the_data, replace = TRUE)`.  

Below we calculate the bootstrap distribution for mean age of OkCupid users using just the first 20 OkCupid users in the data set. 


```{r boot_dist}


# read in the okcupid data
profiles <- read.csv("profiles_revised.csv")



# get the ages from the first 20 OkCupid profiles
ages <- profiles$age

ages_sample <- ages[1:20]


# create the bootstrap distribution

boot_dist <- NULL
for (i in 1:1000) {
  
  boot_sample <- sample(ages_sample, 20, replace = TRUE)
  boot_dist[i] <- mean(boot_sample)
  
}

boot_sample <- sample(ages_sample, 20, replace = TRUE)
mean(boot_sample)


# plot the bootstrap distribution to make sure it looks normal
hist(boot_dist)

(SE_boot <- sd(boot_dist))

(sample_mean <- mean(ages_sample))

CI_upper <- sample_mean + 2 * SE_boot
CI_lower <- sample_mean - 2 * SE_boot

c(CI_lower, CI_upper)

sd(ages_sample)


```



$\\$




#### Part 2.2: Calculating the bootstrap standard error SE*


The standard deviation of the *bootstrap distribution* is usually a good approximation of the standard deviation of the sampling distribution - i.e, it is a good approximation of the *standard error SE*.

When our bootstrap distribution is relatively normal, we can use the fact that 95% of values fall within to standard deviations of a normal distribution to calculate 95% confidence intervals as:

$CI_{95} = [stat - 2 \cdot SE^*,  stat + 2 \cdot SE^*]$

For example, for a our bootstrap distribution we have a 95% confidence interval for the mean $\mu$ as: 

$CI_{95} = [\bar{x} - 2 \cdot SE^*,  \bar{x} + 2 \cdot SE^*]$




```{r bootstrap_SE}

# calculate the bootstrap standard error SE* as the standard deviation of the bootstrap distribution 



# calculate the 95% CI using SE*





```



Above we are using the bootstrap to create a 95% confidence interval which should capture the mean age $\mu_{age}$ 95% of the time.









