---
title: "Class 8 notes and code"
output:
  html_document: default
  pdf_document: default
---





$\\$





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)


options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Permutation tests for more than two means




$\\$











## Part 1: Permutation tests for more than two sample means in R



A group of Hope College students wanted to see if there was an association between a student’s major and the time it takes to complete a small Sudoku-like puzzle

They grouped majors into four categories
 * Applied science (as)
 * Natural science (ns)
 * Social science  (ss)
 * Arts/humanities (ah)





$\\$




##### Step 1: State the null and alternative hypotheses 







$\\$





##### Step 2a: Plot the data


```{r sudoku_plot}


# get the data
sudoku_data <- read.table('http://www.math.hope.edu/isi/data/chap9/MajorPuzzle.txt', header = TRUE)



# create a boxplot of the data



```






$\\$







##### Step 2b: Calculate the observed statistic




```{r sudoku_obs_stat}


# we can use the by() function to get the means separately for each group
# (we will learn of a better way to do this later in the semester)
group_means <- as.vector(by(sudoku_data$time, sudoku_data$major, mean))




# a function to get the MAD statistic
get_MAD_stat <- function(group_means) {
  
  total <- 0
  
  for (iGroup1 in 1:(length(group_means) - 1)) {
    for (iGroup2 in (iGroup1 + 1):(length(group_means))){
      total <- total + abs(group_means[iGroup1] - group_means[iGroup2])
    }
  }
  
  total
  
}  # end of the function 




# apply this function to our data to get our observed statistic 

  

```






$\\$







##### Step 3: Create the null distribution


```{r sudoku_null_dist}


# create the null distribution here using a for loop and the get_MAD_stat() function 






# plot the null distribution as a histogram




```




$\\$






##### Step 4: Calculate a p-value


```{r sudoku_p_value}





# plot the null distribution again with a line at the observed statistic





# calculate the p-value






```





$\\$







##### Step 5: Make a decision






$\\$













