---
title: "Class 20 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()



```




```{r setup, include=FALSE}

# install.packages('car')

library(knitr)
library(latex2exp)
library(dplyr)

#options(scipen=999)

opts_chunk$set(tidy.opts=list(width.cutoff=60)) 
set.seed(230)  # set the random number generator to always give the same random numbers

```



$\\$




## Overview


 * Data exploration and analysis of baseball data
 



$\\$







## Part 1: Multiple linear regression using baseball data




#### Part 1.1: Creating a few commonly used baseball statistics


Let's explore which statistic has the highest correlation with the number of runs a team has scored. 

We can start by creating a few commonly used statistics/metrics for quantifying hitting ability by running the code below. 


```{r, message = FALSE}


# install.packages('Lahman')

library(Lahman)
library(dplyr)


# View(Teams)


# reduce the data frame to the main batting statistics of interest
team_batting <- select(Teams, yearID, teamID, G,  W,  L, R,  AB,  H,  X2B, X3B,  HR,  BB,  SO, SB, CS,  HBP, SF)


# get commonly used composite statistics 
team_batting <- mutate(team_batting,
                       X1B = H - (X2B + X3B + HR), 
                       BA = H/AB, 
                       OBP = (H + BB)/(AB + BB),
                       SlugPct = (X1B + 2 * X2B + 3 * X3B + 4 * HR)/AB,
                       OPS = OBP + SlugPct)


# only use teams that have played the 162 games in a season (the number of games currently in a MLB season)
team_batting <- filter(team_batting, G == 162)

  

```





$\\$






#### Part 1.2: Which statistics are most correlated with runs



Let's look at the coefficient of determination ($R^2$) between runs a team scores (R) and the following statistics:

 * HR:  home runs
 * BA:  batting average
 * SlugPct: slugging percentage
 * OBP:  on base percentage



```{r}


# r^2 for runs with HR
cor(team_batting$R, team_batting$HR)^2


# r^2 for runs with BA
cor(team_batting$R, team_batting$BA)^2


# r^2 for runs with SlugPct
cor(team_batting$R, team_batting$SlugPct)^2


# r^2 for runs with OBP
cor(team_batting$R, team_batting$OBP)^2


# r^2 for runs with OPS
cor(team_batting$R, team_batting$OPS)^2


```




$\\$





#### Part 1.3: Visualizing the relationships



Let's visualize the correlation between the main variables of interest: R, HR, OBP, and SlugPct.



```{r}

pairs(select(team_batting, R, HR, OBP, SlugPct, OPS))


```






$\\$









#### Part 1.4: Comparing David Ortiz and Derek Jeter's OPS



On-base plus Slugging (OPS) seems like the composite statistic that has the highest correlation with runs. Let's compare David Ortiz and Derek Jeter's Slugging percentage to see who is better. 





```{r}

# create Ortiz_info info from the Master data frame 
(Ortiz_info <- Master %>%
  filter(nameFirst == "David", nameLast == "Ortiz"))


# create Jeter_info info from the Master data frame 
(Jeter_info <- Master %>%
    filter(nameFirst == "Derek", nameLast == "Jeter"))


# create Ortiz_Jeter_batting that has Ortiz and Jeter's batting stats
Ortiz_Jeter_batting <- Batting %>%
  filter(playerID %in% c(Ortiz_info$playerID, Jeter_info$playerID))


# add SlugPct to the Ortiz_Jeter_batting data frame
Ortiz_Jeter_batting <- Ortiz_Jeter_batting %>%
  mutate(X1B = H - (X2B + X3B + HR), 
         OBP = (H + BB)/(AB + BB),
         SlugPct = (X1B + 2 * X2B + 3 * X3B + 4 * HR)/AB,
         OPS = OBP + SlugPct)


# create a boxplot to compare them
boxplot(OPS ~ playerID, data = Ortiz_Jeter_batting, names = c("Derek Jeter", "David Ortiz"), 
        ylab = "OPS")


```





$\\$





#### Part 1.6: Fitting a regression model to come up with a better metric


Let's now use the lm() function to fit a linear regression model that can potentially do a better job at predicting runs.

We can also evaluate which variables seem to be useful for predicting runs.




```{r}


# fit the multiple regression model
fit <- lm(R ~ HBP + BB + X1B + X2B + X3B + HR, data = team_batting)


# examine the coefficients
(baseball_coefs <- coef(fit))


# look at the model fit statistics
(summary_fit <- summary(fit))


```



**Questions:**  Do the values for the coefficients make sense? 






$\\$







#### Part 1.7: How good is our new statistic? 

```{r}

# extract the r^2 value
summary_fit$r.squared


```






$\\$





#### Part 1.8: Can we compare David and Derek on our new statistic? 



```{r}


# use the predict() function to get the OPT_stat for Ortiz and Jeter
Ortiz_Jeter_batting$OPT_stat <- predict(fit, newdata = Ortiz_Jeter_batting)


# create a box plot to compare them
boxplot(OPT_stat ~ playerID, 
        data = Ortiz_Jeter_batting, 
        names = c("Derek Jeter", "David Ortiz"), 
        ylab = "OPS")


# run a t-test to see if there is a statistically significant difference between these players
t.test(OPT_stat ~ playerID, data = Ortiz_Jeter_batting)



```






$\\$







#### Part 1.9: Can you come up with a regression model that has an even higher correlation with runs? 


We can add new variables to the data frame that contain additional statistics (e.g., HR^2, etc) and see if you can come up with an even higher correlation value with runs. 


Do you think your new model will be better at predicting the number of runs a team scores in the upcoming 2021 baseball season? 



```{r}





```


**Question:** Did something go wrong here? 





$\\$







#### Part 1.10: Colinearity


Multicolinearity occurs when predictor variables are highly correlated with each other. If there is multicolinearity, there are multiple variables that can make equally good predictions for $y$ and this ends up making the coefficients on these variables not statistically significant. 

The variance inflated factor (VIF) is a statistic to assess multicolinearity. A rule of thumb is that $VIF_i$ > 5 indicates significant multicolinearity. 



```{r}

# library(car)




```






$\\$






