---
title: "Class 8 notes and code"
output:
  pdf_document: default
  html_document: default
---





$\\$





```{r setup, include=FALSE}

# install.packages("latex2exp")

library(latex2exp)


options(scipen=999)


knitr::opts_chunk$set(echo = TRUE)

set.seed(123)

```



$\\$




## Overview

 * Permutation test for two means using a t-statistic
 * Parametric tests for two means




$\\$




## Part 1: Permutation tests for comparing two means in R


Let's us reexamine the randomized comparative experiment by Lyle et al (1987) to assess whether calcium lowered blood pressure.






### Step 1: State the null and alternative hypotheses 








$\\$






### Step 2a: Plot the data


```{r pill_plot}



treat <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1,  12,  -1,  -3,   3,  -5,   5,   2, -11,  -1,  -3)



# plot the data






```





$\\$





### Step 2b: Calculate the observed statistic using a t-statistic


The formula for a t-statistic is: 

$t = \frac{\bar{x}_t - \bar{x}_c}{\sqrt{\frac{s^2_t}{n_t} + \frac{s^2_c}{n_c}}}$




```{r pill_obs_stat}







```


$\\$





### Step 3: Create the null distribution using a permutation/randomization test



```{r null_dist_pill}


# combine the data from the treatment and control groups together



# use a for loop to create shuffled treatment and control groups and shuffled statistics 
null_distribution <-  NULL




















```



$\\$





### Step 4: Calculate a p-value


```{r p_value_pill}


# plot the null distribution with a red line a the value of the observed statistic







# calculate the p-value








```



$\\$



### Step 5: Make a decision








When we used a statistic of $\bar{x}_t - \bar{x}_c$ in our randomization test in class 7 we got a p-value of 0.066. How do these results compare? 







$\\$










## Part 2: Parametric tests for comparing two means in R


Let's redo this analysis using a parametric probability distribution, which in this case is the t-distribution. The same 5 steps of hypothesis testing apply here as well!




$\\$




### Step 1: State the null and alternative hypotheses 


Same as before...








$\\$





### Step 2b: Calculate the observed statistic using a t-statistic


Same as before: 


$t = \frac{\bar{x}_t - \bar{x}_c}{\sqrt{\frac{s^2_t}{n_t} + \frac{s^2_c}{n_c}}}$




```{r pill_obs_stat2}









```




$\\$




### Step 3: Create the null distribution



We will now use a parametric t-distribution (i.e., density function) as a null distribution. The t-distribution has one parameter called "degrees of freedom". We will set this parameter as the minimum of $n_t - 1$ or $n_c - 1$. 


What are the degrees of freedom for this calcium study? 


Let's visualize the t-distribution


```{r t_density}

# get the degrees of freedom




# plot the parametric null distribution




# how does this compare to our t-distribution created by shuffling? 




```


$\\$




### Step 4: Calculate a p-value


We can get $Pr(X > stat)$ for a t-distribution using the `pt()` function.



```{r p_value_pill2}





```





$\\$


### Step 5: Make a decision


How does our p-value and decision compare to the p-value decision we got from the permutation test? 




$\\$




### Built in R functions for t-tests

We can use the built in `t.test()` function to run a t-test as well. We will use the extra argument `alternative = "greater"` to do a one-tailed t-test. 



```{r t_test}





```


Why is the p-value slightly different than what we got when we used the `pt()` function? 








