---
title: "Class 3 notes and code"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




$\\$





# Download files 

If you have not done so already (or if you're working in a new directory), please download files needed for the next two classes. Please press the green play arrow once to download the files. Once they are downloaded you do not need to run this again. 

On the homework, you will also need to start by downloading files also by pressing the green start arrow on an R chuck at the top of the document. 


```{r download_files, echo = FALSE, eval = FALSE}


# If you don't have the SDS230 package working yet, you can download the files using the following commands


download.file("https://raw.githubusercontent.com/emeyers/SDS230/master/ClassMaterial/data/daily_bike_totals.rda", "daily_bike_totals.rda", mode = "wb")

download.file("https://raw.githubusercontent.com/emeyers/SDS230/master/ClassMaterial/data/profiles_revised.csv", "profiles_revised.csv", mode = "wb")


```








$\\$






## Warm up exercise 


The code below loads the OkCupid data 

Can you write code that will print out the mean income of the 2nd person, 4th and 12th person in the data set?


Here are steps you can use:

1. Create a vector called `income` that has everyone's incomes
2. Create a vector called `income3` that has the incomes of the 2nd, 4th and 12th person
3. Take the mean of the `income3` vector. 



```{r warmup}

# Load the data
profiles <- read.csv("profiles_revised.csv")

# View(profiles)


# Create a vector called `incomes` that has everyone's incomes



# Create a vector called `income3` that has the incomes of the 2nd, 4th and 12th person



# Take the mean of the `income3` vector




```




$\\$






## Examining categorical data

Categorical variables take on one of a fixed number of possible values 

For categorical variables, we usually want to view:

 * How many items are each category    or
 * The proportion (or percentage) of items in each category

Let's examine categorical data by looking at drinking behavior on OkCupid


```{r categorical_1}

# Get information about drinking behavior



# Create a table showing how often people drink



```





$\\$





We can create a relative frequency table using the function: `prop.table(my_table)` 

Can you create a relative frequency table for the drinking behavior of the people in the OkCupid data set in the R chunk below? 


```{r categorical_2}




```





$\\$






### Bar plots (no pun intended?)

We can plot the number of items in each category using a bar plot: `barplot(my_table)` 

Can you create a bar plot for the drinking behavior of the people in the OkCupid data set? 

```{r categorical_3}





```





$\\$





Is there a problem with using the bar plot function without any of the extra arguments? 

[XKCD illusterates the point](https://xkcd.com/833/)

Can you figure out how to fix your plot? 

We can also create pie charts using the pie function

```{r categorical_4}




```



Some pie charts [are more informative than others](http://robslink.com/SAS/democd91/pyramid_pie.htm)






$\\$






Our plots are dominated by social drinkers - let's remove them...

```{r categorical_5}








```

[There are other websites/apps for dating as well](https://i.kym-cdn.com/photos/images/original/001/293/992/d80.jpg)





$\\$






## Examining quantiative data

There are several summary statistics useful for describing quantitative data such as the mean and the median. Use the mean() and median() functions to extract measures of the central tendency for OkCupid user's heights.

```{r quantitative_1}




```



What went wrong? 


We can ignore missing values using the `na.rm = TRUE` argument


```{r quantitative_2}




```




Fact: the average height of males in US is 69.6", and of females is 64". Also 60% of the ok cupid users are in our data set are male. Is is the height of the average OkCupid user what we would expect from the US population? Would we expect them to be the same? 



```{r quantitative_3}

expected_okcupid_height <-  .6 * 69.6 + .4 * 64  # based on census data





```






$\\$






We can plot histograms of heights using the `hist()` function. 

```{r quantitative_4}






```




$\\$




We can add lines to our plots using the abline() function. For example abline(v = 60) would add a vertical line at the value of 60. Can you add a vertical line at the average OkCupid user's height? 



```{r quantitative_5}





```




$\\$





Boxplots visually show a version of a 5 number summary (min, Q1, median, Q3, max). We can create boxplots using the `boxplot()` function. 

Create a boxplot of OkCupid user's heights.

```{r quantitative_6}




```

If there are extreme outliers in a plot we need to investigate them. If they are errors we can remove them, otherwise we need to take them into account. 



$\\$





Let's now look at data from CitiBike in New York City. How many cases are there any how many variables? What does each case corresond to? 

```{r quantitative_7}

# download.file("https://yale.box.com/shared/static/t3ezfphfg729x03079aajop0d3f454wm.rda", "daily_bike_totals.rda")

load("daily_bike_totals.rda")

```





$\\$




Scatter plots show the relationships between two quantitative variables. We can use the `plot(x, y)` function to create scatter plots. Create a scatter plot of the maximum temperature as a function of the minimum temperature. Also create a line of the number of trips as a function of the date using the `plot(x, y, type = "o")` function.



```{r quantitative_8}


# scatter plot



# line plot





```






## Challenge exercise 


Can you calculate the mean height of only the men in the data set? 

Steps:
1. Create a vector that has everyone's height called `height`
2. Create a vector that lists the self reported sex of each person called `sex`
2. Create a Boolean vector for just the males can `is_male` that is TRUE an individual's sex is male
3. Use the `is_male` vector to get just the ages of the men and store in a vector called `men_ages`
4. Take the mean of the `men_ages` vector



```{r challenge}

# Load the data
#profiles <- read.csv("profiles_revised.csv")

# View(profiles)


# We can extract the columns of a data frame as vector objects using the $ symbol



# Get the sex of each individual



# Create a Boolean vector that is TRUE if an individual is male
# use == to test if values are equal



# Get the ages of the men



# Get the mean height of men



```




$\\$

