---
title: "Final Project: Add title here"
output:
  pdf_document: default
---



$\\$



## Author:  [Put your name here]

## Discussants: [List people who helped you with the project and/or websites used]







<!--  


This is a template for creating your final project report. It lays out the
sections that should be in your write-up and describes a little about these
sections. There is some flexibility to deviate from this structure, for example,
interweaving more visualizations and analyses could work well.

Your report should be between 5-8 pages long and should contain:

    1. Introduction: 
      a. What is question you are addressing? 
      b. Why is important? 
      c. Where did you get the data?
      d. What other analyses that been done on the data ?
      
    2. Visualizations of the data: one or more plots
    
    3. Analyses: models, hypothesis tests, confidence intervals and other
    inferential statistics that give insight into your question
    
    4. Conclusions: What you found, future directions, etc.
    
    5. Reflection (do be completed on Canvas)
       a. What went well? 
       b. What did you struggle with?
       c. What analyses did you do that you are not including? etc. 

Please make your report look good by paying attention to detail, using
additional R Markdown features etc.

If there is additional code or data you would like to include with your report,
please create a GitHub page and add a link to it in your report. Additionally,
you can append the full code for your analysis in an appendix section at the end
of the document, and then include only the most important pieces of code in the
body of the report. For example, you can exclude details of data cleaning from
the body of the report. However, include anything central to your analyses, and
also any information about particular choices you made that might affect the
results, and why you made those choices, in the body of the report (e.g.,
explain if data was excluded and why, etc.).



--> 









<!-- There are some options that might help make your document look better.  
Feel free to add additional options here -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}

library(knitr)

# This makes sure the code is wrapped to fit when it creates a pdf
opts_chunk$set(tidy.opts=list(width.cutoff=60))   


# Set the random number generator to always give the same random numbers
set.seed(230)  


```







$\\$









## Introduction 


<!--  

Write ~1-3 paragraphs describing:

1. What is problem you are addressing and why the problem you are addressing is
interesting.

2. Where you got the data from, including a link to the website where you got
the data if applicable.

3. What other analyses have already been done with the data and possibly links
to other analyses. Also mention if you are using the data in another class or
for another research project.


--> 





   
$\\$   
   
    




## Results



$\\$
    


### Data wrangling: Change the subtitle here to describe what you are plotting etc.



<!--  

Very briefly discuss how you got the data into shape for your analyses. You can
include some code here, although extensive data cleaning code should be put on
GitHub and/or in an appendix at the end of the document.

--> 




$\\$
    




### Visualize the data: Change the subtitle here to describe what you are plotting etc.



<!--  

Create one or more plots of your data. Describe the plot(s), what they shows,
and why they are of interest to your analysis. Include the code to create these
plots in the R chunk below. You can also discuss the plots after the code too.
Finally, be sure to make plots as clear as possible (clear axis labeling,
legends and captions) so that it is easy for the reader to quickly understand
the central information being conveyed.

--> 




```{r message=FALSE, warning=FALSE, tidy=TRUE}




```




<!--  

Possible additional discussion of the plots here. 

--> 








$\\$    
    







    

### Analyses: Sub-title about the analyses/models you are using 



<!--  

Build linear models, run hypothesis tests, create confidence intervals and/or
run simulations to answer questions that are of interest.

--> 




```{r message=FALSE, warning=FALSE, tidy=TRUE}




```








$\\$


    
    
    
    
    
    


## Conclusion 



<!--  


~1-2 paragraphs summarizing what you found, how the findings address your
question of interest, and possible future directions. Please make sure describe
your conclusions in an intuitive way, and make sure that your argument is strong
and backed by solid evidence from your data.



-->










$\\$






## Reflection


<!--  


Reflection  

Write one paragraph describing what went well with this project and what was
more difficult. Also describe any additional things you tried that you did not
end up including in this write-up, and approximately how much time you spend
working the project.

Finally, please go to Canvas and fill out the Final Project reflection to let use know how the final project (and the class in general) went. 



-->




$\\$




## Appendix


<!--  


You can include a complete listing of your code here if you could not fit it
into the body of the document. Make sure your code is well commented and easy to
read - i.e., use meaningful object names, separate your code into sections,
describe what each section is doing, use good formatting, etc.


-->




