---
title: "Homework 7"
output:
  pdf_document: default
  html_document:
    df_print: paged
---




$\\$





The purpose of this homework is: 1) to examine how influential points can affect regression models, 2) to examine the analysis of variance for regression and 3) to practice building multiple linear regression models. Please fill in the appropriate code and write answers to all questions in the answer sections, then submit a compiled pdf with your answers through Gradescope by 11pm on Sunday November 7th. 

As always, if you need help with the homework, please attend the TA office hours which are listed on Canvas and/or ask questions on [Ed Discussions](https://edstem.org/us/courses/12764/discussion/). Also, if you have completed the homework, please help others out by answering questions on Ed Discussions which will count toward your class participation grade.


**Note: for all plots on this homework, please only use base R graphics**




<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
# SDS230::update_installed_packages()


# Note: I only have permission to use this data for educational purposes so please do not share the data
download.file('https://yale.box.com/shared/static/gzu5lhulepp3zsyxptwxoeafpst1ccdv.rda', 'car_transactions.rda')

```




<!-- The R chunk below sets some parameters that will be used in the rest of the document. Please ignore it and do not modify it. -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo = FALSE}

library(knitr)
library(latex2exp)
library(dplyr)   

options(scipen=999)

opts_chunk$set(tidy.opts=list(width.cutoff=60)) 
set.seed(230)  # set the random number generator to always give the same random numbers
    

```





$\\$





## Part 1: The effect of unusual points on regression models


In class we discussed several ways that data points can be unusual, which are:

1. **Outliers (large residuals)**: data points that have unusual *y* values.
2. **High leverage points**: data points that have unusual *x* values.
3. **Influential points**: data points that are both outliers and have high leverage. 


Let's continue using the Edmunds.com car data to explore statistics that quantify these unusual points, and also examine how unusual points can affect regression models. Remember that this data set has been made available to this class for educational purposes, so please do not share this data outside of the class. 






$\\$







**Part 1.1 (6 points)**: 


Let's start our analysis by again reducing the Edmunds data to only used Toyota Corollas. However, for this analysis, we will create two data sets, one that has all used Corollas and a second that has only Corollas that have been driven less than 150,000 miles. 

To create the data set that has all used Toyota Corollas, please complete the following steps, and save the results in a data frame called `used_corollas_all`:


1. The only variables that should be in the `used_corollas_all` data frame are: 

 a. `model_bought`: the model of the car 
 b. `new_or_used_bought`: whether a car was new or used when it was purchased
 c. `price_bought`: the price the car was purchased for
 d. `mileage_bought`: the number of miles the car had when it was purchased


2. The only cases that should be in the `used_corollas_all` data frame are: 

 a. Used cars
 b. Toyota Corollas

3. Finally use the `na.omit()` function on the `used_corollas` data frame to remove cases that have missing values.


If you have done this correctly, there should be 249 cases in the `used_corollas_all` data frame. Please check this before going on to the next set of exercises. 

Also, please recreate the `used_corollas` data set, that has only used Corollas with less than 150,000 miles (hint: you should be able to do this in 1-2 lines of code). If you look at the `used_corollas` data frame, you will see that there are 248 cases. Thus there is only one car that had been driven over 150,000 miles. We will examine how much of an affect this one car could have on our regression model that contains data on 248 other cars. 



```{r get_all_used_corollas}


# load the data set
load("car_transactions.rda")
















```





$\\$






**Part 1.2 (10 points)**: 

Now fit a linear regression model to the `used_corollas_all` that shows the predicted price as a function of the number of miles driven. Save this model to an object called `lm_fit_all`, print the regression coefficients. Also, create a scatter plot of the price as a function of the number of miles driven, and add a red line to our plot showing the regression line for the model with all used Corollas. 

Next create a model called `lm_fit`, that predicts the price of a car as a function of the number of miles driven using only only cars with less than 150,000 miles (i.e., use the `used_corollas` data frame you created above to recreate the linear model you created on homework 6). Then add a green regression line to the scatter plot you created above that shows the linear model for cars with less than 150,000 miles (i.e., the model where the one high leverage point is not included). 

Finally, make a prediction for the price of a car that has been driven 150,000 miles using both the `lm_fit_all` based on the model using all the data, and the `lm_fit` based on using only cars with less than 150,000 miles. In the answer section, report whether these models seem similar in terms of their regression slopes and predictions.



```{r compare_regression_models}





















```


**Answers**:









$\\$







**Part 1.3 (6 points)**: Now create a 95% confidence interval for the value of the regression slope $\beta_1$ based on the model that does **not** include cars with more than 150,000 miles (i.e., the `lm_fit` model). If we were assuming that this confidence interval is reasonable, would the value for the estimated regression slope found in using all the data (i.e., the $\hat{\beta_1}$ using data in `used_corollas_all` that includes the car with 300,000 miles) be a plausible value for what the true parameter value $\beta_1$ that was found by using the `used_corollas` data set? 


```{r CI_less_than_150k}




```


**Answer** 






$\\$







**Part 1.4 (12 points)**: Now sort the data frame `used_corollas_all` so that the rows are in the order from smallest number of miles driven to the most number of miles driven, and store it again the same object called `used_corollas_all`. Refit the `lm_fit_all` using this sorted data frame (as a sanity check, the coefficients found should be the same as before). Then, recreate the scatter plot based on this sorted `used_corollas_all` data and add to this plot both the 99% confidence intervals for **the regression line** in orange, and the 99% prediction interval in blue (again using this sorted `used_corollas_all`). 

Finally, add a green regression line for the model that does not include cars with over 150,000 miles. In the answer section, describe whether the `lm_fit` regression line falls within the `lm_fit_all` confidence interval, *over the range of mileage where most of the data lies* (hint: creating the plot in the console and zooming in can make this easier to see). Also, describe whether the `lm_fit_all` 99% prediction interval appears to cover the range values where 99% of the observed data lies. 


```{r conf_pred_intervals, warning = FALSE}





































```

**Answer** 








$\\$








**Part 1.5 (12 points)**:  Let's analyze the leverage and Cook's distance for the data points in the `used_corollas_all` data set. Calculate the leverage for the data points in this model (i.e., the hat values) and plot the 20 largest leverage values found using a bar plot. Also plot the residuals as a function of the leverage for each point, and use R's built in plot functions to plot Cook's distance for each data point and the standardized residuals as a function of the leverage for each point. Based on the 'rules of thumb' we discussed in class, **how many points** are considered 'very unusual' for the different measures of: 

a. Cook's distance
b. Standardized residuals 
c. Studentized residuals
d. Leverage

As always, please be sure to show your work below by printing out the answers. 


```{r leverage_analysis}








```



**Answer** 

a.  
b. 
c. 
d. 








$\\$









**Part 1.6 (5 points)**:  Above you fit two models: `lm_fit_all` which contained all the used Corollas and `lm_fit` which did not contain the high leverage car with 300,000 miles driven. Describe below which you think is a better model and why? Also describe any limitation to these models. 



**Answer**  







$\\$







## Part 2: Analysis of variance (ANOVA) for regression


Let's continue to use the Edmunds data to explore the analysis of variance (ANOVA) for regression. For this part of the homework, please use the `lm_fit` model, based on the `used_corollas` data that does **not** include the car that has been driven more than 150,000 miles.



$\\$




**Part 2.1 (5 points)**: 

As you recall from class, the ANOVA breaks down the total variability in a response variable y (SSTotal) as a sum of the variability explained by the model (SSModel) and the sum of the variability not explained by the model (SSResidual). 

To do this decomposition, we can pass a linear model that has been fit (e.g., the `lm_fit` object), to the `anova()` function. Please do this now, and in the answer section report the values for the model sum of squares (SSModel), the residual sum of squares (the SSResidual), and the total sum of squares (SSTotal).


```{r anova_table}








```


**Answers:** 







$\\$





**Part 2.2 (10 points)**:  


**This is a "challenge problem" that you should try to figure out without getting help from the TAs.**


We can also extract the SSTotal, SSModel and SSResidual from the original data and from values stored in the `lm_fit` object. Please use the following equations to calculate the SSTotal, SSModel and SSResidual values directly fom the `lm_fit` object, and report these values:


1. Use the `used_corolla` data frame to calculate the SSTotal using the formula $\sum_{i=1}^n(y_i - \bar{y})^2$. Save the results in an object called `SSTotal` and print the results. 

2. Use the `fitted.values` in the `lm_fit` object to calculate SSModel using the formula $\sum_{i=1}^n(\hat{y}_i - \bar{y})^2$. Save the results in an object called `SSModel` and print the results. 

3. Use the `residuals` in the `lm_fit` object to calculate SSResidual using the formula $\sum_{i=1}^n(y_i - \hat{y}_i)^2$. Save the results in an object called `SSResidual` and print the results.  

To check you have the right answers, look at the values you got in question 2.1 above. Also show that SSTotal = SSModel + SSResidual for the values you calculated. 



```{r SS_decomposition}










```




$\\$





**Part 2.3 (6 points)**: As we discussed in class, the *coefficient of determination*, $r^2$, is equal to the percentage of the variance explained by the linear model, i.e., $r^2 = SSModel/SSTotal$. For simple linear regression, $r^2$ is equal to the correlation coefficient squared, which it is why it is denoted $r^2$. 

Please calculate the correlation coefficient between `mileage_bought` and `price_bought`, and square it. Then using the values calculated in part 2.2, show that this value is equal to SSModel/SSTotal, and that it is also equal to 1 - SSResidual/SSTotal. As always, be sure to show your work by printing these values to get credit for this problem.


```{r coefficient_of_determination}









```






$\\$






**Part 2.4 (5 points)**: Let's look at the relationship between the ANOVA F-statistic and the t-statistic. Use the `summary()` function on the `lm_fit` model to see the value of the t-statistic (as you had previously done in homework 6, part 2.3).  Then, show that the value of `t_stat` squared is (approximately) equal to the F-statistic found with the `anova()` function in part 2.1 above, by printing the $t^2$ value and reporting the $t^2$ value and the F-statistic value in the answer section below.


```{r t_f_comparison}










```

**Answers**:  







$\\$











## Part 3: Multiple linear regression




Let's continue to use the Edmunds.com car data to explore multiple linear regression where we try to predict a response variable $y$, based on several explanatory variables $x_1$, $x_2$, ..., $x_p$. 




$\\$




**Part 3.1 (5 points):** Let's start by using dplyr to derive a new data set from the original Edmunds `car_transactions` data set. Please create this data set in an object called `car_transactions2` that has the following properties: 

1. It contains a new variable called `years_old` which is the difference between the year the car was sold, and the model year of the car. 
2. It only contains *used* cars.

3. It only contains the variables: `price_bought`, `mileage_bought`, `years_old`, and `msrp_bought`.


If you have created this data frame correctly it should have 17,134 cases, so please check this before continuing.


Also, report what is the maximum and minimum value for the variable `years_old`. Does it make sense that the minimum value of `years_old` is negative? Please explain why. Finally, report the price that the least and most expensive used cars sold for. 



```{r car_transformations}















```



**Answer:** 







$\\$







**Part 3.2 (5 points):**  Now use the `pairs()` function to visualize the correlation between all pairs of variables in the `car_transactions2` data frame. Report whether any variable looks like it has a particularly strong linear relationship with `price_bought` and whether it makes sense that there would be a strong relationship between these variables. 


```{r pair_plot}




```


**Answer:** 








$\\$







**Part 3.3 (10 points):**  Next fit a multiple linear regression model predicting the price a car was bought for using the three variables `mileage_bought`, `years_old`, `msrp_bought` and save the linear fit to an object called `lm_cars`. Then use the `summary()` function to get information about the the linear regression model you fit by: a) saving the output of the `summary()` function to an object called `summary_lm_cars`, and b) printing the output so you can see the result. 

Report below the following information:

1. Are all the regression coefficients statistically significant at the $\alpha = 0.05$ level? 

2. Do the signs for the regression coefficients make sense? Explain why. 

3. Report what percentage of the total sum of squares is explained by this model by looking at the values stored in the `summary_lm_cars` object. 



```{r model_summary}










```



**Answers:**

1. 


2.  




3. 





$\\$









## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes link, and clicking on [Reflection on homework 7](https://yale.instructure.com/courses/68751/quizzes/40869).





