---
title: "Homework 9"
output:
  pdf_document: default
  html_document:
    df_print: paged
---




$\\$





The purpose of this homework is to practice fitting logistic regression models,
to learn how to join data frames, and to learn how to create maps. Please fill
in the appropriate code and write answers to all questions in the answer
sections, then submit a compiled pdf with your answers through Gradescope by
11pm on Sunday November 26th.

As always, if you need help with the homework, please attend the TA office hours
which are listed on Canvas and/or ask questions on 
[Ed Discussions](https://edstem.org/us/courses/24940/discussion/). Also, if you have
completed the homework, please help others out by answering questions on Ed
Discussions, which will count toward your class participation grade.




<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}


# Note: I only have permission to use this data for educational purposes so please do not share the data
download.file('https://yale.box.com/shared/static/gzu5lhulepp3zsyxptwxoeafpst1ccdv.rda', 'car_transactions.rda')


SDS230::download_data("condos_may_2022.rda")



```




<!-- The R chunk below sets some parameters that will be used in the rest of the
document. Please ignore it and do not modify it. -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo = FALSE}

library(knitr)
library(latex2exp)
library(dplyr)   
library(ggplot2)

options(scipen=999)

opts_chunk$set(tidy.opts=list(width.cutoff=60)) 
set.seed(230)  # set the random number generator to always give the same random numbers
    

```





$\\$






## Part 1: Logistic regression


As we have discussed in class, logistic regression can be used to estimate the
probability that a response variable `y` belongs to one of two possible
categories. In these exercises, you will learn how to use logistic regression by
fitting models that can give the probability that a car is new or used based on
other predictors, including the car's price.




$\\$





**Part 1.1 (6 points)**: The code below creates a data frame called
`toyota_data` that has sales information on 500 randomly selected new and used
Toyota cars. It also changes the order of the levels on the categorical variable
`new_or_used_bought` which will make the plotting and modeling work better in
these exercises.

To start, create a visualization of the data using ggplot where price is on the
x-axis and whether the car is new or used is on the y-axis. Use a
`geom_jitter()` to plot points that have less overlap, and adjust the amount of
vertical jitters so that there is a clear separation between the new and used
cars (using `position_jitter()` inside the `geom_jitter()` could be useful). Also
set the alpha transparency appropriately to also help deal with over-plotting.


```{r viz_new_used}

load("car_transactions.rda")

# get toyota cars and change the order of the levels of the new_or_used_bought variable
toyota_data <- filter(car_transactions, make_bought == "Toyota") |>
  mutate(new_or_used_bought = factor(as.character(new_or_used_bought, levels = c("U", "N")))) |>
  mutate(new_or_used_bought = relevel(new_or_used_bought, "U")) |>
  na.omit() |>
  group_by(new_or_used_bought) |>
  slice_sample(n = 500)



# visualize the data






  

```





$\\$







**Part 1.2 (8 points)**: Now fit a logistic model to predict the probability a
car is new based on the price that was paid for the car using the `glm()`
function. Then extract the offset and slope coefficients and save them to
objects called `b0` and `b1`. Based on these coefficients, calculate what the
predicted log-odds, odds and probability are that a car is new if it costs
$15,000. Print these values out, and in the answer section, report what these
values are.

A note on R and logistic models: `new_or_used_bought` is a factor variable where
"U" is the lowest level and "N" is the highest level. This means we can plug the
variable into the `glm` function without further transformation, since R will
interpret this dichotomous factor variable as composed of the 1s and 0s that a
logistic regression expects.




```{r logistic_fit}
















```


**Answer:**

The predicted values that a car is new if it cost $15,000 are:

1. log-odds = 

2. odds = 

3. probability = 







$\\$






**Part 1.3 (8 points)**:


Now using the `b0` and `b1` coefficients calculated in part 1.2, write a
function `get_prob_new()` that takes a price a car was sold for as an input
argument and returns the probability the car was new using the equation:

$$\text{P(new|price)} = \dfrac{\exp(\hat{\beta}_0 + \hat{\beta}_1 \cdot x_{price})}{1 + \exp(\hat{\beta}_0 + \hat{\beta}_1 \cdot x_{price})}$$

Then use the `get_prob_new()` function to predict the probability that a car
that sold for $10,000 was a new car and report this probability. Comment on
whether you would expect a predicted probability this high based on the
visualization of the data you created in part 1.1.


```{r logistic_probability_function}

















```

**Answer**: 







$\\$







**Part 1.4 (8 points)**: Next plot the data again as you did in part 1.1, but
this time you will add a red line showing the logistic regression function for
predicting the probability a car is new based on its price. In order to plot the
logistic regression line at the appropriate height on the figure, create a
function called `plot_prob_new(price)` that returns the value of
`get_prob_new(price)` plus 1.

Adding 1 is necessary because R understands the "U" level of our factor variable
to be 1 numerically and "N" to be 2, but `get_prob_new` returns probabilities
between 0 and 1.

Then, use the `stat_function()` function in the ggplot2 package, in combination
with the `plot_prob_new()` function, to create a visualization where you create
the scatter-plot visualization you created in part 1.1. Make sure that you also
overlay the logistic regression line in red on the plot. Use `xlim` to toggle
the dimensions of the x-axis appropiate, such that the plot centers the points
and shows all aspects of the plotted function line.


```{r viz_logistic, warning = FALSE}

















```




$\\$






**Part 1.5 (10 points)**: Let's now fit a multiple logistic regression to the
data in order to get the probability a car is new given the price of the car
(`price_bought`) and the number of miles driven (`mileage_bought`). From the
model that you fit, extract the regression coefficient estimates and intercept,
price_bought and mileage_bought coefficients and save them to objects called
`b0`, `b1` and `b2`. Then write a function called `get_prob_new2()` that uses
the coefficients to predict the probability a car is new using the equation:

$$\text{P(new|price, mileage)} = \dfrac{\exp(\hat{\beta}_0 + \hat{\beta}_1 \cdot price + \hat{\beta}_2 \cdot mileage)}{1 + \exp(\hat{\beta}_0 + \hat{\beta}_1 \cdot price + \hat{\beta}_2 \cdot mileage)}$$

Finally, use the `get_prob_new2()` function to predict the probability that a
car that sold for 10,000 and had 500 miles was a new car, and the probability
that a car that sold for 10,000 and had 5000 miles was a new car. Report these
probabilities in the answer section.



```{r multi_logistic, warning = FALSE}















```


**Answer**: 


For a car that sold for $10,000 and had 500 miles, the predicted probability that the car is new is: 

For a car that sold for $10,000 and had 5000 miles, the predicted probability that the car is new is:  




$\\$







**Part 1.6 (5 points)**:

**This is a "challenge problem" that you should try to figure out without getting help from the TAs.**


We can also fit logistic regression models with categorical predictors. The code below creates a data set that selects 500 random new and used Toyotas and BMWs. Let's use this data to build a model that predicts whether a car is new or used based on the cars price and whether it is a Toyota and BMW. To start, use ggplot to visualize the data where: `price_bought` is mapped to the x-axis, `new_or_used_bought` is mapped to the y-axis, `make_bought` is mapped to color. Also use the `geom_jitter()` glyph using an appropriate amount of jitter, and as always, label your axes. 

Then, fit a logistic regression model to predict whether a car is new or used based on the price when the car was bought and whether the car was a Toyota or BMW. Finally, print out odds ratio for predicting whether a car is new or used if it is a Toyota relative to a BMW. Report in the answer section what the odds ratio is and how to interpret what it means. 


```{r}

set.seed(230)

# get Toyota cars and change the order of the levels of the new_or_used_bought variable
toyota_bmw_data <- filter(car_transactions, make_bought %in% c("Toyota", "BMW")) |>
  mutate(new_or_used_bought = factor(as.character(new_or_used_bought), levels = c("U", "N"))) |>
  mutate(new_or_used_bought = relevel(new_or_used_bought, "U")) |>
  na.omit() |>
  group_by(new_or_used_bought, make_bought) |>
  slice_sample(n = 500) 


















```


**Answer**: 









$\\$




## Part 2: Practice building regression models


To help consolidate what you have learned this semester, and to prepare for the
final project, let's practice building another regression model trying to
predict the price of a condominium in New Haven. In particular, you will go through the full model building process of choosing, fitting, assessing, and using a model.

**Motivation**: In May 2022, I was interested in buying a condominium in New
Haven. One condominium I was interested in was located at 869 Orange St, Unit
5E. The asking price for the condominium was $475,000. However, usually buyers
don't put in offers that are exactly at the asking price, but instead put in
offers above or below. The advantage of putting in a lower offer is that one
could save money, but there is a risk someone else could put in a higher offer
that would be taken instead of your offer. 

In order to help get a sense of how much money I should offer, I scraped data on
all the house prices in New Haven from the New Haven online assessment website
(https://gis.vgsi.com/newhavenct/). The goal of this exercise is to use this
data to create visualizations and regression models in order to come up a offer
that you think will be very likely to be accepted, but that will not be way more
than the property is really worth (i.e., more than you could resell the property
for in the future).

When building regression models below, the response variable you will try to
predict is `sale_price` which is the price that different condominium in New
Haven sold for. Explanatory variables that might be useful to include in your
model include:

1. `zone`: Contains information about the neighborhood where condominiums are
located.

2. `year_built`: The year the condominium was built.

3. `building_area`, `living_area` and `size_acres`: Information about the size of the property. 

4. `appraisal`: How much tax collectors assess that the property was worth.

5. `sale_date`: The year the property was sold. 

You can also use other variables that in the data that are not listed above. For
more information about different variables, see the online assessment website
https://gis.vgsi.com/newhavenct/ (I don't know much more about these variables
than the information that is listed on this site).


Finally, note that there is no one "correct answers" to the exercises below.
Rather, just use visualizations and modeling to come up with "useful" models
that can given insight into the question of interest and we can discuss more
about the models everyone has developed in class.





$\\$





**Part 2.1 (5 points)**: Before building regression models, it is almost always
useful to apply transformations of the data. For example, it could be useful to
filter the data to use only subsets of the data that are most relevant. It could
also be useful to apply transformations to variables in the dataset. For
example, the price a property sold for is going to be heavily influenced on how
long ago the property sold, so converting `sale_date` into a variable such as
`year_old_sold` which indicates how many years ago the property was sold (and
hence the year the is from `sale_price`) could also be useful.

In the R chunk below, please transform the data in anyway you think would be
useful for subsequent analyses and save your transformed data to a data frame
called `property_data` (3-5 transformations/filtering operations should be fine,
but there is flexibility here to do what you think is best).  Then in the answer
section please**briefly** describe the rationale behind any data you removed
(i.e., rationale behind any data filtered out of the dataset).

Hints: the `lubridate` library contains many useful functions for dealing with
dates. For example, the `year()` function can take a date and extract the year
from it, etc.




```{r condo_transformations}

library(dplyr)
library(lubridate)


load("condos_may_2022.rda")










```


**Answer**: 









$\\$






**Part 2.2 (5 points)**: Before modeling the data, one should almost always 
visualize the data. Please go ahead to create 2 visualizations that give insight
into the housing market and/or into how to begin to model the data. In the
answer section briefly describe what your plot(s) are showing and how they given
insight into the question of interest.



```{r condo_viz, message = FALSE}


library(ggplot2)
# library(GGally)











```


**Answer**: 








$\\$







**Part 2.3 (5 points)**:  Now please go ahead and create a regression model for 
predicting the sales price of condos (i.e., the `sale_price` variable) from
other explanatory variables. You can include however many terms in the model
that you think leads to a good model (our model ended up using 5 variables and
an interaction term, but you might be able to come up with a better model so
please do what you think is best).

Also, in the R chunk below, print out a few relevant statistics about the model,
and show appropriate diagnostic plots that give an indication of whether you
model is fitting the data well (again, there is not one "right answer" here but
just use a few of the methods we have discussed).

In the answer section, describe whether you think your model does a reasonable
job fitting the data.

Note, in the process of creating your regression model you might fit several
intermediate models, but only show your final model in the code chunk below.


```{r condo_modeling}






  




```


**Answer** 









$\\$







**Part 2.4 (3 points)**: Now that you've done your analyses, you are ready to 
make an offer on the condominium! 

In the R chunk below, please use the model you created in part 2.3 to make a
prediction for what the price is for the 896 Orange ST 5E condo. Hint: the code
below extracts a data frame with just this relevant condo, and you can use the
`predict()` using the `newdata` argument to get a prediction from your model.


Then in the answer section below please do the following:

1. Write down the predicted price for the property at 869 Orange St Unit 5E
and whether you think this is a reasonable prediction or if you think
it is too high or too low.

2. Write down what you think the best offer would be to put in for trying
to buy this condominium and in 1-3 sentences explain how you came to this
number.  Remember, you definitely want your offer accepted, but you do not want
to over pay so that you are "underwater" on the condominium and you would lose a
lot of money if you had to sell it.

3. Enter these values in the homework 9 reflection and we will examine the
distribution of offers the class comes up with. When you enter your numbers,
please make sure to enter them a single numbers without any commas. For example
if you think the fair price is the asking price of $475,000 then enter the
number 475000 into the homework 3 reflection.




```{r condo_predictions}

# get a data frame with just the 869 ORANGE ST #5E condo
# orange_st_condo <- filter(property_data, street_number == "869 ORANGE ST #5E")









```





**Answer**:

1. 





2. 






$\\$










## Part 3: Thoughts on your final project (3 points)

Describe what you are thinking of doing for your final project and where you
will get your data from. If you are not sure yet what you are going to do for
your final project, that is fine and you can just say that. However, on homework
10 you will need to load the data you will use for your final project data to
demonstrate that you have found relevant data, so please start preparing for
this now. I encourage you to email TA's and ULA's, or to talk to me after class
for guidance.


**Answer**








$\\$







## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes link, and clicking on Reflection on homework 9








