---
title: "Homework 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---




$\\$




# Welcome to the first homework assignment




$\\$



The purpose of this homework is to review concepts from Introductory Statistics and to practice analyzing and plotting data in R. Please fill in the appropriate R code and write answers to all questions in the answer section. Once you have completed the assignment, submit a compiled pdf with your answers to Gradescope by 11:59pm on Sunday September 13th. 

If you need help with the homework, please attend the TA office hours which are listed on Canvas and/or ask questions on [Piazza](https://piazza.com/class/kd52xzes5se3gh). Also, if you have completed the homework, please help others out by answering questions on Piazza which will count toward your class participation grade.



<!-- The R chunk below sets some parameters that will be used in the rest of the document. Please ignore it and do not modify it. -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo = FALSE}

library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60)) 
set.seed(123)  # set the random number generator to always give the same random numbers
    
```




<!--  Please run the code in the  R chunk below once. This will install some packages and download data needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

install.packages("Lahman")

download.file("https://yale.box.com/shared/static/1syezdzked6v2l3aqqpffmifww0468ez.rda", "misc_data.Rda")

download.file("https://yale.box.com/shared/static/p1akv464cut78w859pqqn15ebroxa2wm.rda", "players_born_1901_1950.Rda")

```






$\\$






## Exercise 1: Descriptive statistics and plots for quantitative data


Below you will create and compare a few plots. Please answer each question, and if you notice any outliers in your data please address them appropriately. Also be sure to label your plots appropriately. 



$\\$




**Exercise 1.1: (10 points)** The code chunk below loads four vector objects named x1, x2, x3, and x4. Create a side-by-side boxplot that compares these four vectors. Also create a histogram for each of these vectors (4 histograms total). Describe below whether the box plots or histograms are more informative for plotting this data and why. 

```{r question_1.1}

load("misc_data.Rda")



```

**Answer:** [Describe whether box plots or histograms are more informative here] 



$\\$





**Exercise 1.2: (15 points)** The R chunk below loads a data frame with information on all major league baseball players who were born between 1901 and 1950 (if you are interested in the data, it comes from the Lahman package). Create a vector object that is called `heights`, that has just the player heights. Then create a histogram and a boxplot of the players' heights using this vector object. Describe the shape of the distribution of heights and any advantages that one type of plot has over the other. Also investigate any unusual features of the data.

```{r question_1.2}

load("players_born_1901_1950.Rda")




  
```

**Answer:** [Describe advantages of boxplots and histograms for this data and investigate usual features of the data]. 







$\\$






**Exercise 1.3: (10 points)** Create a scatter plot of the baseball player's heights as a function of their weight. Describe what the results show.

```{r question_1.3}





```

**Answer:**  







$\\$





## Exercise 2: Descriptive statistics and plots for categorical data


Let's now examine which states/regions baseball players are born in.



$\\$


**Exercise 2.1: (10 points)** Use the table() function to create an object called `birth_place_counts` that has the counts of where players were born in. What is the state that the most players were born in and what is the proportion of players born in that state? 

Then create a bar plot and pie chart showing the counts of places that players are born in. How do these plots look? How could we make them better? 


```{r question_2.1}







```


**Answers:** 







$\\$





**Exercise 2.2: (10 points)** 

Let's only plot states/places that have more than 20 players born in them. You can do this by creating a vector of booleans where TRUE indicates a state that has greater than 20 players born in it and FALSE indicates that 20 or less players were born in it (this can be done in 1 line of code). Then use this vector to extract only the places which more than 20 players born in. Finally replot the results with only states with more than 20 players born in them. Does this look better? Is there any place on this list that is not a state? 

```{r question_2.2}






```



**Answer:** 




$\\$





**Exercise 2.3: (10 points)** 

The plots in exercise 2.2 still could look better. Adjust the plots so that you plot fewer states so that it is easier to see exactly which states most players are born in. Also adjust other visual attributes of the plots so that none of the labels are overlapping, and see if you can find other ways to make the plots look better, e.g., by adjusting the colors, etc. (hint: using ? pie and google will be helpful). 


```{r question_2.3}







```



$\\$




**Exercise 2.4: (15 points)** 

The plots made above are a bit misleading because they ignore all players from states in which only a few players were born in. The pie chart is particularly misleading because it gives the sense that a proportion of the players come from a particular state when this proportion is only out of the states that have a large number of players born in them. A way to address this is to create a category called "other state" that has the number of players born in all states that are not shown. 

In the R chunk below, please create a pie chart and a bar plot that have the category "other states" which lists the total number of people born in the other states that were not included in figures you created in exercise 2.3. Again, adjust visual attributes of the plots to make the plots look better. 


Hints: 

1. If you have a vector (or table) called `my_vec`, you can concatenate the value 27 on to the vector `my_vec` using the syntax `my_vec <- c(my_vec, 27)`. 

2. You can get the names of a named vector `my_vec` using the function `names(my_vec)` which returns a vector of strings. You can also assign names to a vector `my_vec` using the function `names(my_vec) <- vec_of_names`, where  `vec_of_names` is a vector of character strings.



```{r question_2.4}














```






## Problem 3: For loops (15 points)


As discussed in class, for loops allow you to repeat a process many times. Each time the process is repeated, a counter index object (usually named *i*) is incremented by 1. This is useful because it allows you to:

1. Repeat a process many times to generate results each time.
2. Store each result in a vector using *i* to index into the vector. 

The code below uses a for loop to store the values of 1 squared up to 50 squared in a vector object named `my_vec`. Modify the code so that what is stored in the vector are even integers from 2 to 100 (i.e, 2, 4, 6, ..., 100).



```{r for_loop_practice}


my_vec <- NULL  # need to initialize the vector so R knows what to put the results into

for (i in 1:50) {
  my_vec[i] <- i^2
}


my_vec


```






$\\$





## Reflection (5 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes link, and clicking on [Reflection on homework 1](https://yale.instructure.com/courses/61201/quizzes/27475)







