---
title: "Homework 10"
output:
  pdf_document: default
  html_document:
    df_print: paged
---




$\\$





The purpose of this homework is to gain practice running analysis of variance hypothesis tests. Please fill in the appropriate code and write answers to all questions in the answer sections, then submit a compiled pdf with your answers through Gradescope by 11:59pm on Sunday November 29th. 

As always, if you need help with any of the homework assignments, please attend the TA office hours which are listed on Canvas and/or ask questions on [Piazza](https://piazza.com/class/kd52xzes5se3gh). Also, if you have completed the homework, please help others out by answering questions on Piazza.



<!--  Please run the code in the  R chunk below once. This will install some packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# makes sure you have all the packages we have used in class installed
SDS230::update_installed_packages()

SDS230::download_data("movies.Rdata")

```




<!-- The R chunk below sets some parameters that will be used in the rest of the document. Please ignore it and do not modify it. -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo = FALSE}

library(knitr)
library(latex2exp)
library(dplyr)   
library(ggplot2)

options(scipen=999)

opts_chunk$set(tidy.opts=list(width.cutoff=60)) 
set.seed(230)  # set the random number generator to always give the same random numbers
    

```





$\\$





 
## Part 1: One-way Analysis of Variance (ANOVA) for testing more than two means



In class 9 we used dplyr to examine which genres of movies critics and regular audiences enjoy the most based on the ratings of over 500 movies from the [Rotten Tomatoes website](https://www.rottentomatoes.com/). In the analyses you did in class 9 you only created descriptive statistics to explore whether different genres of movies had different scores. 

Let's now use a one-way ANOVA to test whether there are indeed differences in the mean of critics scores for different genres of movies. In particular, in this first set of questions you will use the movie data to assess whether there is a difference in mean critic scores for the genres "Action & Adventure", "Comedy" and "Drama".




$\\$




**Part 1.1 (3 points)**:  Start your hypothesis test with step 1, by stating the null and alternative hypotheses in symbols and words. Also state the alpha level that is most commonly used. 



**Answers**


**In words**



**In symbols**



**The significance level**






$\\$





**Part 1.2a (5 points)**:

Now use dplyr to create a data frame called `films`. This data set should be derived from the movies data set and should have the following properties:

1. The type of movie (i.e., `title_type`) should only be "Feature Films".

2. The only variables in the data frame should be: `title`, `genre`, `runtime`, `mpaa_rating`, `critics_score`, and `audience_score`. 

3. The only genres should be "Action & Adventure", "Comedy", and "Drama".

4. The only MPAA ratings should be: "PG", "PG-13" and "R".

5. Be sure to apply the `droplevels()` function to the final data frame to remove all levels of categorical variables that you are not using. 

If you have created this data frame correctly, it should contain 433 rows and 6 columns. 

Once you have created this data frame, create a box plot of the critics scores as a function of genre. Based on this plot, do you believe there is a difference in the average scores critics give to these different movie genres? 


```{r get_rotten_tomatoes_data}


# load the data
load('movies.Rdata')





```

**Answer** 





$\\$






**Part 1.2b (15 points)**: In order for the F-distribution to be a valid null distribution for our F-statistic, two conditions must hold. The two conditions that need to be met are: 1) the variances (or standard deviations) in each group must be approximately the same and 2) the data from each group must be relatively normal (also, as with almost all hypothesis tests, it needs to be the case that the data points are independent, but we will assume that is the case here). One can check these conditions either at the start of the analysis, or at the end before one draws a final conclusion. Let's check these conditions now!

We can check the first condition, as to whether the variances within each group are approximately the same, by comparing the standard deviations between the groups. As long as the largest standard deviation is not twice as big as the smallest standard deviation this condition is met. Please use dplyr to check this condition and in the answer section below describe whether this condition appears to be met. 

We can check the second condition, as to whether the data is relatively normal in each group, by visually inspecting the data. This can This can be done in a several ways including: 1) creating a histogram of the residuals (differences between each point and its group mean), and 2) creating a Quantile-Quantile plot between the the residuals and a normal distribution. 

To create these plots it will be useful to first create a data frame called `films2` that is derived from the `films` data frame but also has two additional variables. The first variable is called `mean_genre_scores` and should have the average critics score for each genre, and the second variable is called `critic_genre_residuals` and should contain the difference of each critics score from the mean genre score. Using the `group_by()` function in conjunction with the `mutate()` function will be helpful for doing this. 

Once you have created the `films2` data frame, create the Q-Q plot using either the `car::qqPlot()` function or the `qqnorm()` function. Also use either ggplot or base R to create a histogram of the residuals (you could also plot a normal distribution on top of this histogram but this is not required). Based on this data, do the residuals appear normally distributed?

Note: as discussed in class, the ANOVA is fairly robust to departures from homoskedasticity and normality. However, if one is concerned about these conditions not being met, one could always run a permutation test instead (see class 15).


```{r check_anova_conditions}










```


**Answer**







$\\$




**Part 1.2c (15 points)**: Next let's use dplyr to create an ANOVA table. ANOVA tables have the following form:

|   Source    |   df    | Sum of Sq.  |   Mean Square    | F-stat  | p-value |
|-------------|---------|-------------|------------------|---------|---------|
|   Groups    |  K - 1  |   SSG       | MSG = SSG/(K-1)  | MSG/MSE |         |
|   Error     |  N - K  |   SSE       | MSE = SSE/(N-K)  |         |         |
|   Total     |  N - 1  |   SSTotal   |                  |         |         |

Where: 

$SSG = \sum_{i = 1}^K n_i (\bar{x}_i - \bar{x}_{tot})^2$
$SSE = \sum_{i = 1}^K\sum_{j = 1}^{n_i}(x_{ij} - \bar{x}_i)^2$
$SST = \sum_{i = 1}^K\sum_{j = 1}^{n_i}(x_{ij} - \bar{x}_{tot})^2$


Please use dplyr to fill in the values in the ANOVA table using the `films2` data frame (using the `films2` will be easier than using the `films` data frame). Be sure to print out all the values as you compute them in the R chunk below to "show your work" (i.e., print out the dfs, SSG, SSE, MSG, etc.). You will fill in the p-value in this table later during part 1.5 of this homework. 

Note: You can use the `lm()` `aov` or `anova()` to check your answers, but **you are not allowed to use these functions** to compute the F-statistic; i.e., you need to use just the `films2` data frame and dplyr to fill in the ANOVA table and to compute the F-statistic. 



```{r movies_obs_stat}








```

**Answer:** 

|   Source    |   df    | Sum of Sq.  |   Mean Square    | F-stat  | p-value    |
|-------------|---------|-------------|------------------|---------|------------|
|   Groups    |         |             |                  |         |            |
|   Error     |         |             |                  |         |            |
|   Total     |         |             |                  |         |            |





$\\$






**Exercise 1.3 (6 points)**: Now it's time to do step 3 of the hypothesis test by plotting the null distribution. To create the appropriate F-distribution, use the degrees of freedom you calculated in part 1.2c. Then use either base R graphics or ggplot to plot the F-distribution density function, and add the observed statistic as a red vertical line to the plot. From looking at this null distribution, what do you think the p-value is?


```{r F_null_distribution}







```



**Answer**







$\\$







**Part 1.4 (4 points)**: Now do step 4 of hypothesis testing by calculating the p-value using the `pf()` function. Report what the p-value is (and make sure you look at the correct tail). Is this close to what you estimated by looking at the null distribution above? Also, fill in the p-value in the ANOVA table in part 1.2c above.

```{r p_value}


```


**Answers** 






$\\$








**Part 1.5 (3 points)**:  Now complete step 5 of hypothesis testing by making a judgment. Are you able to reject the null hypothesis? What do you conclude? 


**Answer** 





$\\$





**Exercise 1.6 (5 points)**: As we have discussed, we can use R's `lm()` function to run an ANOVA.  Running an ANOVA and creating table requires two steps: 

1. We must fit a model using the syntax: `fit <- lm(response_variable ~ categorical_predictor, data = my_data)`

2. We can then print an ANOVA table using `anova(fit)`

Please use the `lm()` function to fit a model that predicts critics score from the different genres of movies. Save the model that you fit to an object called `fit_genre`. Then use the `anova()` function to see whether the results match the results you had in parts 1.1 to 1.5 above. Report below whether the results match.


```{r r_diagnositcs}




```

**Answer**:  





$\\$





**Exercise 1.7 (5 points)**: 

When we use the `lm()` function to run an ANOVA, we can use the `plot()` function on the model we have fit to get diagnostic plots which we can use to assess whether assumptions underlying the ANOVA were met (i.e., whether the residuals are normally distributed with equal variance). Please use the `plot()` function on the `fit_genre` model you created above to create diagnostic plots. Briefly discuss if you think these plots are more or less informative relative to the plots you created in part 1.2b. Also describe what is determining the x-axis location of the points in the residuals vs. fitted values plot (upper left plot), and why all the points are clustered together on the x-axis. 



```{r}

# create 4 subplot with diagnostic plots




```

**Answer**:







$\\$
 




## Part 2: Two-way analysis of variance 


Let's now run a two-way ANOVA to assess whether both genre and the MPAA rating a movie gets affects critics' scores. 



$\\$



**Part 2.1 (12 points)**:  Start your hypothesis test with step 1 by stating the null and alternative hypotheses in symbols and words. State these null and alternative hypotheses for the main effect of genre, the main effect of MPAA rating, and also for the interaction effect of genre and MPAA rating. 



#### Main effect for genre


**In words**




**In symbols**




#### Main effect for MPAA rating


**In words**



**In symbols**




#### Interaction effect for genre and MPAA rating 


**In words**



**In symbols**







$\\$






**Part 2.2 (5 points)**: 

Now fit a model **that only has main effects** for genre and MPAA rating (i.e., that does not contain an interaction term). Use the `Anova()` function in the `car` package to create an ANOVA table using a type III sum of squares. Are the main effects statistically significant? 



```{r two_way_main_effects}




```

**Answer** 





$\\$





#### Part 2.3 (7 points): Interaction effects

Now let's visualize the data to see if there is an interaction between genre and MPAA rating. Use the `interaction.plot()` function to visualize the film's genre on the x-axis with different lines for the different MPAA rating levels. Try to make the plot look nice by making sure the labels are meaningful and choose a decent color scheme. Based on the this visualization, does there seem to be an interaction between genre and MPAA rating?


```{r interaction_plot}



```

**Answer**: 






$\\$






#### Part 2.4 (points): Testing interactions effects


Now fit a model has both the main effects and an interaction effect for genre and MPAA rating. Again use the `Anova()` function in the `car` package to create an ANOVA table using a type III sum of squares. Are the effects for all factor levels statistically significant? 



```{r two_way_interaction}




```

**Answer**: 





$\\$





#### Part 2.5 (7 points): Checking ANOVA assumptions


In order for our inferences to be valid, the assumptions underlying the ANOVA should be met. Please check these assumptions now and report whether they appear to be met. 



```{r two_way_assumptions}

par(mfrow = c(2, 2))






```

**Answer**





$\\$





## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes link, and clicking on [Reflection on homework 10](https://yale.instructure.com/courses/61201/quizzes/27486).








