---
title: "Homework 7"
output:
  pdf_document: default
  html_document:
    df_print: paged
---




$\\$





The purpose of this homework is: 1) to examine how influential points can affect
regression models, 2) to examine the analysis of variance for regression and 3)
to practice building multiple linear regression models. Please fill in the
appropriate code and write answers to all questions in the answer sections, then
submit a compiled pdf with your answers through Gradescope by 11pm on Sunday
November 3rd.

As always, if you need help with the homework, please attend the TA office hours
which are listed on Canvas and/or ask questions on Ed Discussions. Also, if you
have completed the homework, please help others out by answering questions on Ed
Discussions, which will count toward your class participation grade.


**Note: for all plots on this homework, please only use base R graphics**




<!--  Please run the code in the  R chunk below once. This will install some
packages and download data and images needed for these exercises.  -->

```{r message=FALSE, warning = FALSE, echo = FALSE, eval = FALSE}

# Note: I only have permission to use this data for educational purposes so please do not share the data
download.file('https://yale.box.com/shared/static/gzu5lhulepp3zsyxptwxoeafpst1ccdv.rda', 'car_transactions.rda')

```




<!-- The R chunk below sets some parameters that will be used in the rest of the document. Please ignore it and do not modify it. -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo = FALSE}

library(knitr)
library(latex2exp)
library(dplyr)   

options(scipen=999)

opts_chunk$set(tidy.opts=list(width.cutoff=60)) 
set.seed(230)  # set the random number generator to always give the same random numbers
    

```





$\\$





## Part 1: The effect of unusual points on regression models


In class we discussed several ways that data points can be unusual, which are:

1. **Outliers (large residuals)**: data points that have unusual *y* values.  
2. **High leverage points**: data points that have unusual *x* values.  
3. **Influential points**: data points that are both outliers and have high leverage.  


Let's continue using the Edmunds.com car data to explore statistics that
quantify these unusual points, and also examine how unusual points can affect
regression models. Remember that this data set has been made available to this
class for educational purposes, so please do not share this data outside of the
class.






$\\$







**Part 1.1 (6 points)**: 

Let's start our analysis by again reducing the Edmunds data to only used Toyota
Corollas. However, for this analysis, we will create two data sets:

1. `used_corollas_all`: that has for all used Corollas.  
2. `used_corollas`: that has only used Corollas that have been driven less than 150,000 miles.  


To create the data set that has all used Toyota Corollas, please complete the
following steps, and save the results in a data frame called `used_corollas_all`:

1. The only variables that should be in the `used_corollas_all` data frame are: 

 a. `model_bought`: the model of the car 
 b. `new_or_used_bought`: whether a car was new or used when it was purchased
 c. `price_bought`: the price the car was purchased for
 d. `mileage_bought`: the number of miles the car had when it was purchased


2. The only cases that should be in the `used_corollas_all` data frame are: 

 a. Used cars
 b. Toyota Corollas


3. Finally use the `na.omit()` function on the `used_corollas` data frame to
remove cases that have missing values.


If you have done this correctly, there should be 249 cases in the
`used_corollas_all` data frame. Please print out the number of rows using the `
dim()`to check this is correct before going on to the next set of exercises.


Also, please recreate the `used_corollas` data set that has only used Corollas
with less than 150,000 miles (hint: you should be able to do this in 1-2 lines
of code). If you use the `dim()` function on the `used_corollas` data frame, you 
will see that there are 248 cases. Thus there is only one car that had been
driven over 150,000 miles. We will examine how much of an affect this one car
could have on our regression model that contains data on 248 other cars.



```{r get_all_used_corollas}


# load the data set
load("car_transactions.rda")
















```





$\\$






**Part 1.2 (12 points)**: 


Now compare the regression model that is fit all the data (`lm_fit_all`), to the
regression that is fit to the data without the 150k mileage car (`lm_fit`). To
do this please complete the following steps:

1. Fit a linear regression model to the `used_corollas_all` data that shows the
predicted price as a function of the number of miles driven. Save this model to
an object called `lm_fit_all` and print out the regression coefficients. 

2. Create a scatter plot of the price as a function of the number of miles
driven using all the data.

3. Add a red line to the plot that shows the regression line for the model with all
used Corollas to this scatter plot. 

4. Next create a model called `lm_fit`, that predicts the price of a car as a
function of the number of miles driven using only only cars with less than
150,000 miles (i.e., use the `used_corollas` data frame you created above to
recreate the linear model you created on homework 6). 

5. Then add a green regression line to the scatter plot you created above that
shows the linear model for cars with less than 150,000 miles (i.e., the model
where the one high leverage point is not included).

6. Finally, print out the predicted value for the price of a car that has been
driven 150,000 miles using both the `lm_fit_all` based on the model using all
the data, and the `lm_fit` based on using only cars with less than 150,000
miles.

7. In the answer section, report whether these models seem similar in terms of
their regression slopes and predictions.


**Hint**: For your linear regression models with `lm()`, make sure to use the syntax
`lm(variable2 ~ variable1, data = df)` rather than `lm(df$variable2 ~ df$variable1)`. 
This is important for getting `predict()` to return only a single value when 
you're using the model to predict a single point, and when using `lm()` to do 
multiple regression.


```{r compare_regression_models}


# fit a regression model to all the data and save it to the object lm_fit_all
# (note: we fit the model by specifying y as as function of x)





# display the regression coefficients




# plot the data as a scatter plot of the price as a function of the mileage driven






# add the regression line to the plot showing the fit from using all the data




# fit a model using the data that excluded the car that had been drive over 150k miles




# add a green line to the scatter plot of the regression line from the model that 
# excludes cars with over 150k miles





# make a prediction for a car driven 150k miles using both models









```


**Answers**:









$\\$







**Part 1.3 (6 points)**: Now create a 95% confidence interval for the value of
the regression slope $\beta_1$ based on the model that does **not** include cars
with more than 150,000 miles (i.e., the `lm_fit` model). 

If we were assuming that this confidence interval is accurate, would the value
that was estimated as regression slope using all the data (i.e., the
$\hat{\beta_1}$ based on using the data in `used_corollas_all` that includes the
car with 300,000 miles) be a plausible value for what the true parameter value
$\beta_1$?


```{r CI_less_than_150k}




```


**Answer** 






$\\$







**Part 1.4 (12 points)**: 


Now let's create confidence intervals and predictions intervals
*based on the model that was fit to all the data* and see if the regression
line from model **that does not include the 150k mileage car** falls in
these confidence intervals. To do this, please complete the following steps:

1. Sort the data frame `used_corollas_all` so that the rows are in the order
from smallest number of miles driven to the most number of miles driven, and
store it again the same object called `used_corollas_all`. Refit the
`lm_fit_all` using this sorted data frame (as a sanity check, the coefficients
found should be the same as before).

2. Create and print out the confidence interval values for the regression
coefficients from this model. (i.e., the confidence intervals for $\beta_0$ and
$\beta_1$)

3. Recreate the scatter plot based on this sorted `used_corollas_all` data.

4. Add to this plot the 99% confidence intervals for **the regression line**
in orange (using the model from the sorted `used_corollas_all` data).

5. Add to this plot the 99% **prediction intervals** in blue (again using the
model from the sorted `used_corollas_all` data).

6. Finally, add to this plot the regression line from the model that does
**not** include the car with over 150,000 miles as a green line.

7. In the answer section, describe whether the `lm_fit` regression line falls
within the `lm_fit_all` confidence interval,
*over the range of mileage where most of the data lies* (hint: creating the plot 
in the console and zooming in can make this easier to see). 

8. Also in the answer section, describe whether the `lm_fit_all` 99% prediction
interval appears to cover the range values where 99% of the observed data lies.


```{r conf_pred_intervals, warning = FALSE}


# arrange the data from all the cars and refit the model



# create confidence intervals for the regression coefficients (betas) from the model using all the data




# create confidence interval for the regression line mu_y from the model that used all the data



# create a prediction interval for the regression model that used all the data




# create a scatter plot of all the data





# add the confidence intervals (in orange) for the regression line you created above to the scatter plot




# plot prediction intervals (in blue) you created above to the scatter plot





# add the regression line (in green) based on the model that only uses data from cars with less than 150,000 miles





```

**Answer** 








$\\$









**Part 1.5 (12 points)**:  Let's analyze the leverage and Cook's distance for
the data points in the `used_corollas_all` data set. To do this please: 

1. Calculate the leverage for the data points in this model (i.e., the hat
values). 

2. Plot the 20 largest leverage values as a bar plot (remember all plots in the
homework should use base R graphics).

3. Plot the residuals as a function of the leverage for each point (use just the
regular residuals, not the standardized or studentized residuals).

4. Use R's built in `plot()` functions to plot Cook's distance for each data point.

5. Also, use R's built in `plot()` functions to plot the standardized residuals
as a function of the leverage for each point.

6. Based on the 'rules of thumb' we discussed in class, in the answer section report 
**how many points** are considered 'very unusual' for the different measures of:

a. Cook's distance   
b. Standardized residuals    
c. Studentized residuals  
d. Leverage  

As always, please be sure to also "show your work" by printing out the number of
points that are considered very unusual for each of these measures.



```{r leverage_analysis}


# get the h_i "hat values" and plot the 20 largest values




# plot residuals as a function of the hat values





# plot Cook's distances for each data point 



# plot the standarized residuals as a function of the leverage



# report number of highly unusual points based on Cook's distance



# report number of highly unusual points based on leverage 



# report number of highly unusual points based on standardized residuals



# report number of highly unusual points based on studentized residuals



```



**Answer** 

a.   
b.  
c.  
d.  








$\\$









**Part 1.6 (5 points)**:  Above you fit two models: `lm_fit_all` which contained
all the used Corollas and `lm_fit` which did not contain the high leverage car
with 300,000 miles driven. Describe below which you think is a better model and
why? Also describe any limitation to these models. 

Note, there is not necessarily one correct answer here, but just make sure your
answer is reasonable.



**Answer**  







$\\$







## Part 2: Analysis of variance (ANOVA) for regression


Let's continue to use the Edmunds data to explore the analysis of variance
(ANOVA) for regression. For this part of the homework, please use the `lm_fit`
model, based on the `used_corollas` data that does **not** include the car that
has been driven more than 150,000 miles.



$\\$




**Part 2.1 (5 points)**: 

As you recall from class, the ANOVA breaks down the total variability in a
response variable y (SSTotal) as a sum of the variability explained by the model
(SSModel) and the sum of the variability not explained by the model
(SSResidual).

To do this decomposition, we can pass a linear model that has been fit (e.g.,
the `lm_fit` object), to the `anova()` function. Please do this now, and in the
answer section report the values for the model sum of squares (SSModel), the
residual sum of squares (the SSResidual), and the total sum of squares
(SSTotal).


```{r anova_table}








```


**Answers:** 







$\\$





**Part 2.2 (5 points)**:  



We can also extract the SSTotal, SSModel and SSResidual from the original data
and from values stored in the `lm_fit` object. Please use the following
equations to calculate the SSTotal, SSModel and SSResidual values directly fom
the `lm_fit` object, and report these values:


1. Use the `used_corolla` data frame to calculate the SSTotal using the formula
$\sum_{i=1}^n(y_i - \bar{y})^2$. Save the results in an object called `SSTotal`
and print the results.

2. Use the `fitted.values` in the `lm_fit` object to calculate SSModel using the
formula $\sum_{i=1}^n(\hat{y}_i - \bar{y})^2$. Save the results in an object
called `SSModel` and print the results.

3. Use the `residuals` in the `lm_fit` object to calculate SSResidual using the
formula $\sum_{i=1}^n(y_i - \hat{y}_i)^2$. Save the results in an object called
`SSResidual` and print the results.

To check you have the right answers, look at the values you got in question 2.1
above. Also show that SSTotal = SSModel + SSResidual for the values you
calculated.



```{r SS_decomposition}










```




$\\$





**Part 2.3 (6 points)**: As we discussed in class, the *coefficient of determination*, 
$r^2$, is equal to the percentage of the variance explained by
the linear model, i.e., $r^2 = SSModel/SSTotal$. For simple linear regression,
$r^2$ is equal to the correlation coefficient squared, which it is why it is
denoted $r^2$.

Please calculate the correlation coefficient between `mileage_bought` and
`price_bought`, and square it (hint, the `cor()` function could be helpful).
Then using the values calculated in part 2.2, show that this value is equal to
SSModel/SSTotal, and that it is also equal to 1 - SSResidual/SSTotal. As always,
be sure to show your work by printing these values to get credit for this
problem.


```{r coefficient_of_determination}









```






$\\$






**Part 2.4 (5 points)**: Let's look at the relationship between the ANOVA
F-statistic and the t-statistic. Use the `summary()` function on the `lm_fit`
model to see the value of the t-statistic (as you had previously done in
homework 6, part 2.3).  Then, show that the value of `t_stat` squared is
(approximately) equal to the F-statistic found with the `anova()` function in
part 2.1 above, by printing the $t^2$ value and reporting the $t^2$ value and
the F-statistic value in the answer section below.


```{r t_f_comparison}










```

**Answers**:  







$\\$











## Part 3: Multiple linear regression




Let's continue to use the Edmunds.com car data to explore multiple linear
regression where we try to predict a response variable $y$, based on several
explanatory variables $x_1$, $x_2$, ..., $x_p$.




$\\$





**Part 3.1 (5 points):** Let's start by using dplyr to derive a new data set
from the original Edmunds `car_transactions` data set. Please create this data
set in an object called `car_transactions2` that has the following properties:

1. Create a new variable called `years_old` which is the difference between the
year the car was sold, and the model year of the car.

2. Reduce the data to only coming from *used* cars.  

3. Reduce the data so it only contains the variables: `price_bought`,
`mileage_bought`, `years_old`, and `msrp_bought`.

If you have created this data frame correctly it should have 17,134 cases, so
please check this before continuing (for the sake of simplicity, do not use
`na.omit()` here).

Also, report what is the maximum and minimum value for the variable `years_old`.
Does it make sense that the minimum value of `years_old` is negative? Please see
if you can come up with an explanation why. Finally, report the price that the
least and most expensive used cars sold for.



```{r car_transformations}















```



**Answer:** 







$\\$







**Part 3.2 (5 points):**  Now use the `pairs()` function to visualize the
correlation between all pairs of variables in the `car_transactions2` data
frame. Report whether any variable looks like it has a particularly strong
linear relationship with `price_bought` and whether it makes sense that there
would be a strong relationship between these variables.


```{r pair_plot}




```


**Answer:** 








$\\$







**Part 3.3 (10 points):**  Next fit a multiple linear regression model
predicting the price a car was bought for using the three variables
`mileage_bought`, `years_old`, `msrp_bought` and save the linear fit to an
object called `lm_cars`. Then use the `summary()` function to get information
about the the linear regression model you fit by: a) saving the output of the
`summary()` function to an object called `summary_lm_cars`, and b) printing the
output so you can see the result.

Report below the following information:

1. Are all the regression coefficients statistically significant at the $\alpha = 0.05$ level? 

2. Do the signs for the regression coefficients make sense? Explain why. 

3. Report what percentage of the total sum of squares is explained by this model
by looking at the values stored in the `summary_lm_cars` object.



```{r model_summary}










```



**Answers:**

1. 


2.  




3. 





$\\$









## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes
link, and clicking on Reflection on homework 7.





